### 2019-11-15 dmtn-tbd.tex
initialized

\documentclass[DM,lsstdraft,authoryear,toc]{lsstdoc}
% lsstdoc documentation: https://lsst-texmf.lsst.io/lsstdoc.html

% Package imports go here.
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
\usepackage{enumitem}

% Local commands go here.

% To add a short-form title:
% \title[Short title]{Title}
\title[Alerts Menu]{LSST Alerts: A Menu of Options}

% Optional subtitle
% \setDocSubtitle{A subtitle}

\author{%
M.~L.~Graham, and the Data Management System Science Team
}

\setDocRef{DMTN-MENU}
\date{\today}
% Optional: name of the document's curator
% \setDocCurator{The Curator of this Document}
\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-xxx}}

\setDocAbstract{%
A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
}

\begin{document}

% Create the title page.
% Table of contents is added automatically with the "toc" class option.
\maketitle

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Introduction} \label{sec:intro}

The LSST Science Requirements Document \citedsp{LPM-17} specifies that information about the detections of transient, variable, and moving objects be released promptly as a data stream.

The purpose of this document is to consider the science motivation for, and impact of, the various options for alert packet contents and distribution timescales.

The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public (exempt from a proprietary period) and publicly \emph{accessible}. While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp[LDO-13].

The size and latency of alerts might effect the number of supported community alert brokers. It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout. This is based in part on estimates of the alert stream data rate, and the bandwidth allocated to alert distribution in the LSST data facility.

This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
\begin{itemize}
\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
\item DM-20298, Revise the contents of the alert packets (Graham)
\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
\end{itemize}

% \lsrreq \ossreq \dmreq \reqparam

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Alert Packet Contents} \label{sec:packets}

The section explores the science impacts of various options for alert packet contents, such as the $\sim$12 month history of {\tt DIASource} records, the image stamps, and the static-sky {\tt Object} catalog associations.

One of the positive scientific impacts of reducing the size of an alert packet is based on the assumption that the ability to transmit the stream to more community brokers would lead to more science being done with alerts. There is a potential secondary science impact, in that if brokers can spend less time developing software or spend less of their computational resources to, e.g., remove parts of the alert packet they do not need prior to processing, then they can spend more on developing and running algorithms for various alerts-related science goals. \S~\ref{ssec:packets_remove} and \ref{ssec:packets_compress} consider reducing the size of the alerts.

\subsection{Removing Histories and/or Stamps}\label{ssec:packets_remove}

The size of a fully-loaded individual alert packet is estimated to be $\lesssim$82~KB, based on simulations of the planned content of the alerts as described in Section 3.5 of \citeds{LSE-163}. The two largest components of the alert packet which could be considered "optional" are the history (the past $\sim$12 months of {\tt DIASource} records) and the stamps (at least 30$\times$30 pixels and contain flux (32 bit/pix), variance (32 bit/pix), and mask (16 bit/pix) extensions for both the template and difference image, plus a header of metadata \citedsp{LSE-163}). The history accounts for $\sim$27KB of the alert packet ($\sim$33\%) and the stamps contribute $\gtrsim$18~KB ($\sim$20\%).

Brokers that save all alerts, build a database of alerts contents, or have an automated interface with the LSST Prompt products database or Alerts database, would not need the historical records to be included. Brokers which do not use the image stamps -- or would need them for only a subset of the alerts and could feasibly query a stamps database -- would not need them included in the alerts packets. Individual broker teams may indicate which information they require (or would like removed from the packets in their stream) during the broker proposal process \citedsp{LDM-682}, and \textcolor{red}{we should include that feedback in this document}.

If stamps are not sent in alerts, they need to be stored in a publicly accessible database

{\bf Remove histories and stamps from all alerts. --}

Pro: all transmitted streams have the same content and are identical (no need to keep on hand multiple versions of alert packets).

Con: negatively impact science with brokers which do not save or associate alerts but run in pure real-time (\textcolor{red}{check LOI to assess how many?})

{\bf Remove histories and stamps for some alert streams, at the request of brokers. --}

{\bf Never remove histories or stamps from any distributed alerts. --}

Pro: all transmitted streams have the same content and are identical (no need to keep on hand multiple versions of alert packets).

Pro: every alert packet contains all the public data and there is no need to provide, e.g., a \emph{publicly accessible} stamps database.

\subsection{Adding Associated {\tt Object} Record Elements}\label{ssec:packets_add}

The contents of the alert packet are defined in \citeds{LSE-613}. They will include {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}, {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds, and {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}. For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.

First, this definition of {\tt nearbyObjLnP} seems to only be appropriate for stars. For galaxies, the likelihood of a nearby galaxy being the true host is proportional to {\tt nearbyObjDist} divided by the effective radius of the host, or even better, \textcolor{red}{find the definition that uses semi-major and minor axies and the position angle etc.}. For this reason, the {\tt nearbyObj} included in the alerts should not simply be the three nearest galaxies -- it should be the \emph{three most likely host galaxies}.

Second, the most useful piece of information about a transient's host galaxy is the redshift, because this significantly improves photometric classification algorithms \textcolor{red}{(could cite RAPID or Ashely's work and others)}.

Including the three nearest stars remains appropriate because all stars have the same radial profile. For variable stars, \textcolor{red}{the most useful piece of information about its {\tt Object} is probably the mean apparent magnitude?}

\subsection{Compressing the Alert Packets}\label{ssec:packets_compress}

The application of gzip compression could further reduce the size of a full alert to $\sim$65~KB (80\%; JIRA ticket DM-16280). Naively, this seems like it would allow 1 more full stream to be transmitted to a broker, but the time and computational resources required to compress the alert packets has not been considered, \emph{and} this option would then force all brokers to decompress the alerts on arrival which might further delay processing for everyone. Science goals requiring low-latency alerts (\S~\ref{sec:latency}) could be negatively impacted.

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Alert Distribution Latency} \label{sec:latency}

It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.

Here we attempt to outline which science goals require 1 minute alerts, and could not be satisfied with a longer latency (e.g., 5--10 minutes).

The fast things seem to be triggers in other wavelengths... is there an optical thing you optically detect and then get followup? These scenarios start with a coincident LSST observation with a FRB or BNS GW event...

\subsection{Fast Radio Bursts}

(Can grab all the stuff you put into the CBW paper.)

\subsection{Binary Neutron Star Mergers}
% From Slack conversation

hi Fed, for what concerns ToO observations, I would say a very low latency would be optimal: I know that jet-powered (http://arxiv.org/abs/1705.10797) and/or proto-magnetar-powered (http://arxiv.org/abs/1508.07939) UV (and X-ray) transients from binary neutron star mergers are predicted to be detectable only up to ~1000 s following the merger. I cannot think of such a time-critical case for transients discovered during the main survey though...
arXiv.orgarXiv.org
The cocoon emission - an electromagnetic counterpart to...
Short Gamma-Ray Bursts (SGRBs) are believed to arise from compact binary mergers (either neutron star-neutron star or black hole-neutron star). If so their jets must penetrate outflows that are...
arXiv.orgarXiv.org
Electromagnetic emission from long-lived binary neutron star...
Recent observations indicate that in a large fraction of binary neutron star (BNS) mergers a long-lived neutron star (NS) may be formed rather than a black hole. Unambiguous electromagnetic (EM)...
3 replies

federica bianco  4 months ago
Thanks! Excellent though even in "very low latency" cases, is the difference between 60 sec and 5 min critical?

Eric Bellm  4 months ago
It?s not clear to me whether the 60 second latency could be achieved for rapid TOO observations in any case: in the standard survey we expect to have to do a fair amount of pre-caching of slow database queries to get alerts out in time, which is possible because the next pointings are known ahead of time.  Potentially for a rapid TOO there wouldn?t be time for those to be executed before the images are taken, which would impose some further delay.  We don?t yet know how large an issue this would be, though. (edited) 
:+1:
1

Om Sharan Salafia  4 months ago
@federica I think you're right, it is probably anyway not critical either

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Pre-Filtered Alert Streams} \label{sec:prefilter}

Pre-filtered streams might mean the alerts can be distributed to more brokers than numStreams. What kind of pre-filters might suit the brokers?

 - magnitude or signal-to-noise ratio (affects completeness/purity)
 - region of sky (like the south african proposal)
 - static-sky object association

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Graceful Degradation} \label{sec:graceful}

Along with the requirement of a 60 second alert distribution timescale (\S~\ref{sec:latency}, the DMS should support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 10,000 alerts per standard visit on average during a given night, and at least 40,000 alerts per single standard visit.

It is also a requirement\reqparam{sciVisitAlertDelay}\reqparam{sciVisitAlertFailure}\ossreq{0112}\dmreq{0392} that no more than 1\% of all standard visits fail to have at least 98\% of its alerts distributed within 60 seconds of image readout, and that no more than 0.1\% of all standard visits fail to distribute alerts.

These requirements apply to standard visits which should have produced $\leq$40,000 alerts. For example, a visit would be considered "delayed" and count towards that 1\% limit if $>$2\% of its alerts were distributed with a latency of $>$60 seconds. The requirement that no more than 0.1\% of all science visits fail to generate and/or distribute alerts is integrated over all stages of data handling, not just alert distribution, and includes failures at any stage of prompt processing.

It is furthermore specified that alert distribution "degrade gracefully" beyond that limit, meaning that visits resulting in an excess of alerts should not cause any DMS downtime \citedsp{LSE-30,LSE-61}.

In crowded fields there might be $>$40,000 difference-image sources.

For alerts that are not distributed within 60 seconds, how should they be handled?

It is a requirement that all alerts be stored in an archival database and be available for retrieval\ossreq{0185}. The term "available for retrieval" applies to users with data rights and access to the LSST Science Platform. Like all other Prompt data products, the alerts archive will be updated within 24\reqparam{L1PublicT}\lsrreq{0104} hours \citedsp{LSE-29}.

The retention period in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.

Brokers with data rights can retrieve old alert packets from the Alerts Database. However, note that the requirements on user access to the Alert Database are not yet fully developed, and may be limited in terms of how alert packets may be queried (e.g., alert ID), or by the bulk download capabilities (DMTN-102 1).

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{LSST Alert Filtering Service}\label{sec:lafs}

It is a requirement that the LSST alert filtering service be able to support\reqparam{numBrokerUsers}\dmreq{0343} at least 100 simultaneous users. The LAFS will be capable of returning 20\reqparam{numBrokerAlerts}\dmreq{0343} full-sized alerts per visit per user.

The LSST alerts filtering service is a mechanism by which users --- individuals with LSST data rights and access --- can receive alerts via pre-defined filters that have been optimized for established transient classifications such as supernovae and/or create and apply their own filters to the stream \citedsp{LPM-17,LSE-61}.

Which options might be appropriate for implementation with the LAFS?
A few of these options might also be available to users of the LSST alert filtering service (\S~\ref{sec:LAFS}).

% Include all the relevant bib files.
% https://lsst-texmf.lsst.io/lsstdoc.html#bibliographies
\bibliography{local,lsst,lsst-dm,refs_ads,refs,books}

\end{document}

### 2019-11-15 dmtn-tbd.tex
added draft state warning

@@
 \setDocAbstract{%
+\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT READ. DO NOT CITE.}\\
 A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.

### 2019-11-18 dmtn-tbd.tex
minor updates, work proceeds

@@
 \setDocAbstract{%
-\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT READ. DO NOT CITE.}\\
+\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ (YET!).}\\
 A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
@@
 The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public (exempt from a proprietary period) and publicly \emph{accessible}.
-While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp[LDO-13].
+While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp{LDO-13}.
 
-The size and latency of alerts might effect the number of supported community alert brokers.
 It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout.
-This is based in part on estimates of the alert stream data rate, and the bandwidth allocated to alert distribution in the LSST data facility.
+This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
+The size and latency of alerts could affect the ultimate number of supported community alert brokers.
 
@@
 
-Here we attempt to outline which science goals require 1 minute alerts, and could not be satisfied with a longer latency (e.g., 5--10 minutes).
+This section attempts to gather the science goals which might potentially require 1 minute alerts, and could not be satisfied with a longer latency (e.g., 5--10 minutes).
 
-The fast things seem to be triggers in other wavelengths...
-is there an optical thing you optically detect and then get followup?
-These scenarios start with a coincident LSST observation with a FRB or BNS GW event...
+\subsection{Results of a TVS Survey About Alerts Timescales}
+
+Compact object mergers (kilonovae) appeared 4 times as the response to a question about the science driver for rapid alerts, but each respondent chose a different "maximum tolerable" time delay for alerts: of 1 min or less, 5 to 10 min, 1 to 6 hours, and 12 to 24 hours.
+The respondent's chosen "ideal average" times are only a little more consistent: 1 min or less (x2), 10-30 min, and 30-60 min.
+This probably represents diverse science goals to be done with kilonovae ?
+I?ll look into the details some more?
+
+\subsection{Solar System Objects}
 
@@
 \subsection{Binary Neutron Star Mergers}
-% From Slack conversation
+% From Slack conversation, Federica Bianco, Om Sharan Salafia, Eric Bellm
+
+During LSST Operations, target-of-opportunity imaging follow-up sequences might be executed in the error ellipse of a gravitational wave (GW) detection to search for the electromagnetic (EM) optical counterpart.
+Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
 
@@
 
-federica bianco  4 months ago
-Thanks! Excellent though even in "very low latency" cases, is the difference between 60 sec and 5 min critical?
-
-Eric Bellm  4 months ago
-It?s not clear to me whether the 60 second latency could be achieved for rapid TOO observations in any case: in the standard survey we expect to have to do a fair amount of pre-caching of slow database queries to get alerts out in time, which is possible because the next pointings are known ahead of time.
-Potentially for a rapid TOO there wouldn?t be time for those to be executed before the images are taken, which would impose some further delay.
-We don?t yet know how large an issue this would be, though.
-(edited)
-:+1:
-1
-
-Om Sharan Salafia  4 months ago
-@federica I think you're right, it is probably anyway not critical either
+{\bf Summary --} Alert distribution within 5 minutes of readout is likely sufficient for all science goals related to BSN mergers and their associated optical counterparts.
+This is potentially good because for rapid ToO observations with LSST, there might not be time for the pre-caching of slow database queries (e.g., loading slices of the PPDB catalogs) when the upcoming pointings are not known in advance, as they are for the regular survey observations.
+\textcolor{red}{$\leftarrow$ This statement was modified from something ECB mentioned on Slack, double check it's accurate}.
 
@@
  - magnitude or signal-to-noise ratio (affects completeness/purity)
- - region of sky (like the south african proposal)
+ - region of sky (like one of the LOI for a particular region)
  - static-sky object association
+
+ Image free streams of alert packets.
+ Sources only associated with past alerts.
+
+Provide estimates for how much the alert stream volume would be reduced.
 
@@
 
-The retention period in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.
-
+\textcolor{red}{From the community post brokers FAQ $\rightarrow$}The retention period in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.
 Brokers with data rights can retrieve old alert packets from the Alerts Database.

### 2019-11-19 dmtn-tbd.tex
just saving monday's updates

@@
 The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public (exempt from a proprietary period) and publicly \emph{accessible}.
-While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp{LDO-13}.
+While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp{LDO-013}.
 
@@
 
+\clearpage
 % % % % % % % % % % % % % % % % % % % % % % % % % %
@@
 
-The contents of the alert packet are defined in \citeds{LSE-613}.
+The contents of the alert packet are defined in \citeds{LSE-163}.
 They will include {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}, {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds, and {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}.
@@
 
-\subsection{Results of a TVS Survey About Alerts Timescales}
+{\bf Target of Opportunity --} For rapid ToO observations with LSST, there might not be time for the pre-caching of slow database queries (e.g., loading slices of the PPDB catalogs) when the upcoming pointings are not known in advance, as they are for the regular survey observations.
+\textcolor{red}{$\leftarrow$ This statement was modified from something ECB mentioned on Slack, double check it's accurate}.
 
-Compact object mergers (kilonovae) appeared 4 times as the response to a question about the science driver for rapid alerts, but each respondent chose a different "maximum tolerable" time delay for alerts: of 1 min or less, 5 to 10 min, 1 to 6 hours, and 12 to 24 hours.
-The respondent's chosen "ideal average" times are only a little more consistent: 1 min or less (x2), 10-30 min, and 30-60 min.
-This probably represents diverse science goals to be done with kilonovae ?
-I?ll look into the details some more?
+\subsection{Results of a TVS Survey About Alerts Timescales}\label{ssec:latency_tvs}
 
-\subsection{Solar System Objects}
+The Transients and Variable Stars (TVS) science collaboration surveyed the science needs of their members with respect to alert latency, and the results are shown in Table \ref{tab:tvs}\footnote{Results courtesy of TVS co-chair Rachel Street.}.
+The survey asked respondents to choose the maximum tolerable and ideal average delay between the alerts being produced by the LSST data reduction pipeline and the alert information becoming available through the broker service.
+This is not exactly the same as the alert distribution timescale {\tt OTT1}, but these responses will inform the need for $<$1 minute alert distribution.
+Respondents were asked to provide a short summary of their science goals for alerts if they reported needed access within 15 minutes.
+The pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.
 
-\subsection{Fast Radio Bursts}
+Only 20\% (10\%) of respondents report that $<$10 minutes ($\leq$1 minute) is an ideal average delay, and whereas 70\% report that $>$30 minutes would be a sufficient average latency.
+The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
+It might at first be confusing that EM-GW was provided as the science driver for such a wide range of alert access timescales, but this is discussed further in \S~\ref{ssec:latency_emgw}.
 
-(Can grab all the stuff you put into the CBW paper.)
+\begin{table}[h]%[htdp]
+\caption{Table of results from a TVS survey which asked "how fast do you really need alerts?".
+The total number of respondents was 20.
+The science driver acronyms are: EM-GW (electromagnetic counterparts to gravitational wave events), YSNe (young supernovae), SBOs (???), GRBs (gamma-ray bursts).
+\label{tab:tvs}}
+\begin{center}
+\begin{tabular}{|l|cl|cl|}
+\hline
+             & Maximum & Science & Ideal       & Science \\
+Latency & Tolerable  & Driver(s) &  Average & Driver(s) \\
+\hline
+1 min or less & 1 & EM-GW, GRB  & 2 & EM-GW, YSNe, GRB \\
+1-5 min         & 0 &                          & 1 &                                   \\
+5-10 min       & 1 & EM-GW, YSNe & 1 & SBOs \\
+10-30 min     & 2 & SBOs               & 2 & EM-GW, GRBs \\
+30-60 min     & 2 &                          & 7 & EM-GW  \\
+1-6 hours     & 4 & EM-GW, GRBs & 1 &  \\
+6-12 hours   & 2 &                          & 1 &  \\
+12-24 hours & 3 & EM-GW            & 1 & EM-GW \\
+1-3 days      & 1 &                           & 0 &  \\
+>3 days       & 4 &                           & 4 &  \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
 
-\subsection{Binary Neutron Star Mergers}
+\subsection{Solar System Objects}\label{ssec:latency_sso}
+
+\textcolor{red}{MLG: in Lynne's talk on the SSSC needs for alerts brokers at the CBW in June 2019, slide 4 says that a needed broker capability was "API or other 'push' notification (e.g., trailed detections) [minutes may count]".
+Follow-up on the SS use case for alerts on very short timescales.}
+
+\subsection{Young Supernovae}\label{ssec:latency_ysne}
+
+\subsection{Fast Radio Bursts}\label{ssec:latency_frb}
+
+\textcolor{red}{MLG: Can grab all the stuff you put into the CBW paper and put it here instead.}
+
+\subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
 % From Slack conversation, Federica Bianco, Om Sharan Salafia, Eric Bellm
@@
 Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
+
+The timescale required for alerts access depends on the science goal regarding "kilonovae".
+The light curves last only a few days, and there is so far only the one GW170817, and so even if there was a day or two delay between discovery and the onset of optical follow-up, that's still going to lead to great science.
+However, the point of this section is to discuss the science motivation for very rapid alerts, 1 minute vs.
+5 minutes vs.
+10 minutes.
 
@@
 
-{\bf Summary --} Alert distribution within 5 minutes of readout is likely sufficient for all science goals related to BSN mergers and their associated optical counterparts.
-This is potentially good because for rapid ToO observations with LSST, there might not be time for the pre-caching of slow database queries (e.g., loading slices of the PPDB catalogs) when the upcoming pointings are not known in advance, as they are for the regular survey observations.
-\textcolor{red}{$\leftarrow$ This statement was modified from something ECB mentioned on Slack, double check it's accurate}.
+\subsection{Gamma-Ray Bursts}\label{ssec:latency_grb}
 
+\subsection{Summary}\label{ssec:latency_summary}
+
+\clearpage
 % % % % % % % % % % % % % % % % % % % % % % % % % %

### 2019-11-22 dmtn-tbd.tex
just saving monday's updates
Fri Nov 22 updates

@@
 
-\setDocRef{DMTN-MENU}
+\setDocRef{DMTN-TBD}
 \date{\today}
@@
 
-This section attempts to gather the science goals which might potentially require 1 minute alerts, and could not be satisfied with a longer latency (e.g., 5--10 minutes).
+If it is possible to relax this requirement to, e.g., 5 minutes -- \emph{without risk to the science goals} -- then it might be possible to support more community brokers and/or repurpose the human and computational resources that would otherwise be dedicated to delivering 1 minute alert distribution.
+This section attempts to gather the science goals which might potentially require 1 minute alert distribution and could not be satisfied with a longer latency.
 
-{\bf Target of Opportunity --} For rapid ToO observations with LSST, there might not be time for the pre-caching of slow database queries (e.g., loading slices of the PPDB catalogs) when the upcoming pointings are not known in advance, as they are for the regular survey observations.
+{\bf Generally, science goals that require alerts on timescales shorter than $\sim$5 minutes have that need because they also require follow-up on timescales shorter than $\sim$15--30 minutes.} Most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps).
+Thus, any science goals driving the need for alerts on timescales shorter than $\sim$5 minutes \emph{must also be able to confidently identify their targets with single-epoch single-filter LSST photometry}, or be using additional information (such as host galaxy characteristics, or coincidence with another survey's data).
+
+{\bf Target of Opportunity --} It is important to note that for rapid ToO observations with LSST (if they occur), there might not be time to pre-cache slow database queries (e.g., loading slices of the PPDB catalogs) when upcoming pointings are not known in advance, as would happen during regular survey observations.
+This might mean that a 1 minute alert distribution timescale for alerts from ToO imaging surveys cannot be guaranteed, and this should be kept in mind when reading through the science cases for short alert latencies in this section.
 \textcolor{red}{$\leftarrow$ This statement was modified from something ECB mentioned on Slack, double check it's accurate}.
 
-\subsection{Results of a TVS Survey About Alerts Timescales}\label{ssec:latency_tvs}
+\subsection{TVS Survey Results Regarding Alerts Timescales}\label{ssec:latency_tvs}
 
@@
 Respondents were asked to provide a short summary of their science goals for alerts if they reported needed access within 15 minutes.
-The pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.
+Note that the pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.
 
@@
 The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
-It might at first be confusing that EM-GW was provided as the science driver for such a wide range of alert access timescales, but this is discussed further in \S~\ref{ssec:latency_emgw}.
+It is at first confusing that EM-GW was provided as the science driver for such a wide range of alert access timescales, but this is discussed further in \S~\ref{ssec:latency_emgw}, along with several other science cases that rely on rapid access to LSST alerts.
 
@@
 The total number of respondents was 20.
-The science driver acronyms are: EM-GW (electromagnetic counterparts to gravitational wave events), YSNe (young supernovae), SBOs (???), GRBs (gamma-ray bursts).
+The science driver acronyms are: EM-GW (electromagnetic counterparts to gravitational wave events), YSNe (young supernovae, including e.g., shock breakouts), GRBs (gamma-ray bursts).
 \label{tab:tvs}}
@@
 1-5 min         & 0 &                          & 1 &                                   \\
-5-10 min       & 1 & EM-GW, YSNe & 1 & SBOs \\
-10-30 min     & 2 & SBOs               & 2 & EM-GW, GRBs \\
+5-10 min       & 1 & EM-GW, YSNe & 1 & YSNe \\
+10-30 min     & 2 & YSNe               & 2 & EM-GW, GRBs \\
 30-60 min     & 2 &                          & 7 & EM-GW  \\
@@
 
-\subsection{Solar System Objects}\label{ssec:latency_sso}
+\subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
+% From Slack conversation, Federica Bianco, Om Sharan Salafia, Eric Bellm
 
-\textcolor{red}{MLG: in Lynne's talk on the SSSC needs for alerts brokers at the CBW in June 2019, slide 4 says that a needed broker capability was "API or other 'push' notification (e.g., trailed detections) [minutes may count]".
-Follow-up on the SS use case for alerts on very short timescales.}
+During LSST Operations, target-of-opportunity imaging follow-up sequences might be executed in the error ellipse of a gravitational wave (GW) detection to search for the electromagnetic (EM) optical counterpart.
+Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
+Although the kilonova light curves last only a few days, since there is so far only one event, GW170817, a day or two delay between the GW event and the optical detection still yields very scientifically valuable data -- for now.
+During LSST Operations there will already exist a sizable collection of longer-latency follow-up, and it is likely that science will be moving in the direction of pushing to ever earlier detections.
+However, since KNe produce days-long optical afterglows, a 5 minute LSST alerts would likely suffice.
+
+There are two theoretical predictions for prompt optical emission that would require very rapid access to alerts.
+One of them is a potential faint, $<$1 hour, UV/optical transient that occurs at the time of jet-break out for a short gamma-ray burst associated with a binary neutron star merger (\S~\ref{ssec:latency_grb}).
+Another that predicts emission of a similar color, luminosity, and timescale is the spin-down energy of a long-lived ($10^2$--$10^4$ s) neutron star formed from a binary neutron star merger before its eventual collapse to a black hole \citep{2016ApJ...819...15S}.
+As described in \S~\ref{ssec:latency_grb}, very few events could be detected by serendipitous coincidence by the LSST WFD main survey, and targeted follow-up of well-localized GW events with more appropriate facilities (such as space-based UV/optical imagers) is much more likely to yield detections of this very short lived emission.
+
+There are two additional issues related to ToO for EMGW events which might make 1 minute alerts necessary or impossible.
+First, as described above, for a ToO imaging survey with LSST there might not be time to pre-load catalogs for the targeted fields.
+The second is that, at least for run O3, the GW event detection system itself issues preliminary alerts within 1--10 minutes, and these preliminary alerts are often retracted and do not always have the sky localization.
+Even now, many imaging follow-up surveys wait for the initial alert (or retraction) to be sent after a round of human vetting of the GW event signal, and this can take several hours.
+With such latencies, the 1 {\it vs.} 5 minute LSST alert timescales becomes inconsequential.
+
+\subsection{Gamma-Ray Bursts}\label{ssec:latency_grb}
+
+For short gamma-ray bursts (sGRBs) which are thought to be the mergers of two neutron stars, or a neutron star and a black hole, as the jet propagates through the ejecta material a hot cocoon is formed.
+When cocoon and jet break out, along with sGRB can can be observed as a blue transient with an absolute peak brightness of -12 to -15 magnitudes that lasts for $10^3$--$10^4$ seconds (e.g., \citealt{2018MNRAS.473..576G}).
+For an LSST detection limit of $r\sim24$ mag, a detection limit of $r\lesssim-12$ ($r\lesssim-15$) mag corresponds to distances of $\lesssim160$ ($\lesssim630$) Mpc, or redshifts $z\lesssim0.035$ ($z\lesssim0.14$).
+Observing the diversity of cocoon emission will require multi-band follow-up within $10^3$ seconds, or $15$ minutes, and in this particular case a 5 minute alert distribution latency is too long to include the shorter events (or the fainter events for which only the peak would be visible).
+
+% z=0.035 is V=0.014 Gpc3
+% z=0.14 is V=0.836 Gpc3
+% (0.836/0.014) * 4/yr = 240/yr
+
+The rate of short GRBs in the local volume ($<$200 Mpc) is estimated to be quite low, $<$4 $\rm yr^{-1}$ \citep{2019arXiv190800100M}.
+Scaling this up to the co-moving volume within 630 Mpc is $\sim$240 $\rm yr^{-1}$, all-sky.
+In the baseline main survey, about a sixth of the sky can be observed in a given night, which is $\lesssim$0.1 event in the WFD main survey area per night.
+With $\sim$1000 visits per night, and $\sim$10 visits every 5 minutes, there's a $10^{-3}$ chance that the location of a sGRB will be serendipitously observed within 5 minutes of the event on a given night.
+So this could happen once every 3 years, or $\sim$3 times over the 10-year LSST main survey.
+
+For this particular science case, LSST seems to be the wrong tool for the job.
+As described by \citet{2018MNRAS.473..576G}, a rapid search at the location of sGRBs with a UV satellite such as ULTRASAT would be ideal.
+
+%Triggering criteria might be:
+%new {\it u}- or {\it g}-band source with recent limits associated with a $<$200 Mpc galaxy (how many expected per night?)
+%and then very high priority if it's serendipidously associated with a GRB detection in the last 5 minutes
 
@@
 
-\subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
-% From Slack conversation, Federica Bianco, Om Sharan Salafia, Eric Bellm
+\subsection{Solar System Objects}\label{ssec:latency_sso}
 
-During LSST Operations, target-of-opportunity imaging follow-up sequences might be executed in the error ellipse of a gravitational wave (GW) detection to search for the electromagnetic (EM) optical counterpart.
-Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
-
-The timescale required for alerts access depends on the science goal regarding "kilonovae".
-The light curves last only a few days, and there is so far only the one GW170817, and so even if there was a day or two delay between discovery and the onset of optical follow-up, that's still going to lead to great science.
-However, the point of this section is to discuss the science motivation for very rapid alerts, 1 minute vs.
-5 minutes vs.
-10 minutes.
-
-hi Fed, for what concerns ToO observations, I would say a very low latency would be optimal: I know that jet-powered (http://arxiv.org/abs/1705.10797) and/or proto-magnetar-powered (http://arxiv.org/abs/1508.07939) UV (and X-ray) transients from binary neutron star mergers are predicted to be detectable only up to ~1000 s following the merger.
-I cannot think of such a time-critical case for transients discovered during the main survey though...
-arXiv.orgarXiv.org
-The cocoon emission - an electromagnetic counterpart to...
-Short Gamma-Ray Bursts (SGRBs) are believed to arise from compact binary mergers (either neutron star-neutron star or black hole-neutron star). If so their jets must penetrate outflows that are...
-arXiv.orgarXiv.org
-Electromagnetic emission from long-lived binary neutron star...
-Recent observations indicate that in a large fraction of binary neutron star (BNS) mergers a long-lived neutron star (NS) may be formed rather than a black hole. Unambiguous electromagnetic (EM)...
-3 replies
-
-\subsection{Gamma-Ray Bursts}\label{ssec:latency_grb}
+\textcolor{red}{MLG: in Lynne's talk on the SSSC needs for alerts brokers at the CBW in June 2019, slide 4 says that a needed broker capability was "API or other 'push' notification (e.g., trailed detections) [minutes may count]".
+Follow-up on the SS use case for alerts on very short timescales.}
 

### 2020-01-04 dmtn-tbd.tex
first draft S2 completed

@@
 % \title[Short title]{Title}
-\title[Alerts Menu]{LSST Alerts: A Menu of Options}
+\title[Alerts Menu]{LSST Alerts: Science-Driven Options for Packets and their Distribution}
 
@@
 \author{%
-M.~L.~Graham, and the Data Management System Science Team
+M.~L.~Graham, T. Pritchard, and the Data Management System Science Team
 }
@@
 \setDocAbstract{%
-\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ (YET!).}\\
+\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
 A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
@@
 
-The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public (exempt from a proprietary period) and publicly \emph{accessible}.
-While the contents of the Prompt products database (PPDB, from which alerts are generated) are also world public, \emph{access} to the PPDB is restricted to individuals with data rights \citedsp{LDO-013}.
+The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public --- they are exempt from any proprietary period --- and publicly accessible -- they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
+In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
 
@@
 This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
-The size and latency of alerts could affect the ultimate number of supported community alert brokers.
+The size and latency of alerts could therefore affect the ultimate number of supported community alert brokers.
 
@@
 The section explores the science impacts of various options for alert packet contents, such as the $\sim$12 month history of {\tt DIASource} records, the image stamps, and the static-sky {\tt Object} catalog associations.
-
-One of the positive scientific impacts of reducing the size of an alert packet is based on the assumption that the ability to transmit the stream to more community brokers would lead to more science being done with alerts.
-There is a potential secondary science impact, in that if brokers can spend less time developing software or spend less of their computational resources to, e.g., remove parts of the alert packet they do not need prior to processing, then they can spend more on developing and running algorithms for various alerts-related science goals.
-\S~\ref{ssec:packets_remove} and \ref{ssec:packets_compress} consider reducing the size of the alerts.
+Reducing the size of an alert packet might be scientifically beneficial if it enables the stream to be transmitted to more community brokers, thereby leading to more science being done with alerts.
+It might also be beneficial if it enables brokers to spend less time developing software or spend less of their computational resources to remove parts of the alert packet they do not need prior to processing, and spend more on developing and running algorithms for various alerts-related science goals.
+\S~\ref{ssec:packets_remove} and \ref{ssec:packets_compress} consider reducing the size of alert packets by relocating some data or compressing the packets, respectively. 
+A small increase to the size of the alert packet is not necessarily going to be scientifically detrimental if the information added enables science, and \S~\ref{ssec:packets_add} considers some small additions regarding the association of {\tt DIASources} with the data release {\tt Object} catalog.
 
@@
 
-The size of a fully-loaded individual alert packet is estimated to be $\lesssim$82~KB, based on simulations of the planned content of the alerts as described in Section 3.5 of \citeds{LSE-163}.
-The two largest components of the alert packet which could be considered "optional" are the history (the past $\sim$12 months of {\tt DIASource} records) and the stamps (at least 30$\times$30 pixels and contain flux (32 bit/pix), variance (32 bit/pix), and mask (16 bit/pix) extensions for both the template and difference image, plus a header of metadata \citedsp{LSE-163}).
+The size of a fully-loaded individual alert packet is estimated to be $\lesssim$82~KB, based on simulations of the planned content of the alerts as described in Section 3.5 of \citeds{LSE-163} (see also \citeds{DMTN-102}).
+The two largest components of the alert packet which could be considered ``optional'' are the history and the postage stamps.
+The history is the past $\sim$12 months of {\tt DIASource} records, previously released as alerts.
+The postage stamps are, at minimum, 30$\times$30 pixels and contain flux, variance, and mask extensions for both the template and difference image, plus a header of metadata \citedsp{LSE-163}.
 The history accounts for $\sim$27KB of the alert packet ($\sim$33\%) and the stamps contribute $\gtrsim$18~KB ($\sim$20\%).
 
-Brokers that save all alerts, build a database of alerts contents, or have an automated interface with the LSST Prompt products database or Alerts database, would not need the historical records to be included.
-Brokers which do not use the image stamps -- or would need them for only a subset of the alerts and could feasibly query a stamps database -- would not need them included in the alerts packets.
-Individual broker teams may indicate which information they require (or would like removed from the packets in their stream) during the broker proposal process \citedsp{LDM-682}, and \textcolor{red}{we should include that feedback in this document}.
+Brokers that save all alerts, build their own a database of alerts contents, or have an automated interface with the LSST Prompt products database or Alerts database might not need the historical records to be included in the alert packets.
+Brokers which do not use the image stamps -- or would need them for only a subset of the alerts and could feasibly query a stamps database -- might not need them included in the alerts packets.
+Individual broker teams may indicate which information they require (or would like removed from the packets in their stream) during the broker proposal process \citedsp{LDM-612,LDM-682,LDM-723}.
+During the first stages of the proposal process, and at the LSST Community Brokers Workshop\footnote{\url{ls.st/cbw}} in June 2019, there was some indication that at least a few brokers were interested in alerts without historical records and/or postage historical records.
 
-If stamps are not sent in alerts, they need to be stored in a publicly accessible database
+\subsubsection{{\tt DIASource} Record History}\label{sssec:packets_remove_hist}
 
-{\bf Remove histories and stamps from all alerts. --}
+Including the {\tt DIASource} record history in every alert allows for the \emph{instant assessment} --- without the need to cross-match or query catalogs --- of changes in the objects's location, size, shape, or brightness, provides context for the latest detection (i.e., anomalous or typical), and enables a robust assessment of the likely physical nature of the object (e.g., light curve fits indicating supernova type become more reliable with more observations).
 
-Pro: all transmitted streams have the same content and are identical (no need to keep on hand multiple versions of alert packets).
+To remove the {\tt DIASource} record history from an alert would slightly inhibit the scientific assessment of transient/variable objects for brokers that build their own database of alert contents, due to the time required to cross-match with the database.
+Removing the {\tt DIASource} record history would completely inhibit this kind of analysis for brokers that do not have such storage or processing capabilities.
+There is also the issue that the association of {\tt DIASources} into {\tt DIAObjects} may change over time, for some objects.
+For example, rare cases such as multiply-imaged strongly-lensed supernovae, or two unassociated transients/variables that are superimposed along the line of sight, might be blended/separated in poor/good seeing.
+In such cases, the set of all associated {\tt DIASources} might be improved between one alert and the next.
 
-Con: negatively impact science with brokers which do not save or associate alerts but run in pure real-time (\textcolor{red}{check LOI to assess how many?})
+{\bf Summary --} The {\tt DIASource} record histories should not be removed from all alert packets, but could be optionally removed at the request of a broker.
+However, if the record histories are removed, a list of all {\tt DIASource} identifiers (the {\tt diaSourceId}, as in Section 3.3.1 of \citeds{LSE-163}) which are currently associated with the {\tt DIAObject} to which the alert-spawning {\tt DIASource} is associated should be provided (note that this list is \emph{not} part of the {\tt DIAObject} record).
 
-{\bf Remove histories and stamps for some alert streams, at the request of brokers. --}
+\subsubsection{Forced Photometry History}\label{sssec:packets_remove_fp}
 
-{\bf Never remove histories or stamps from any distributed alerts. --}
+When forced photometry ({\tt DIAForcedSource} record history) is included in an alert packet, it is ``historical" in the sense that it is based on past images and not the image which spawned the alert.
+However, it is not always redundant information that was previously released in an alert, like the {\tt DIASource} record history.
+There are two main cases where forced photometry is performed, as described in \citeds{LSE-163}.
+(1) For all new {\tt DIAObjects}, forced ``pre-covery'' photometry is done in the past 30 days of difference images within 24 hours, stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
+(2) For all {\tt DIAObjects} that are not detected\footnote{The detection threshold is a signal-to-noise ratio $\geq$ 5.} in a given difference image, but had a detection in the past $\sim$12 months, forced photometry is done on that difference image.
+The result is stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
+The forced photometry is scientifically valuable because, for example, pre-covery forced photometry provides context for the first detection of a transient/variable, and low-SNR detections can help with photometric classification and prioritization for follow-up observations.
 
-Pro: all transmitted streams have the same content and are identical (no need to keep on hand multiple versions of alert packets).
+If a given {\tt DIAObject} is never again detected in a difference image, then the {\tt DIAForceSource} records are never included in an alert packet, but are available to individuals or brokers with access to the PPDB via the LSST Science Platform.
+If a given {\tt DIAObject} is repeatedly detected in new difference images, then the same {\tt DIAForcedSource} records will be redundantly included in all future alerts (after the first one containing the pre-covery or non-detection forced photometry).
+For the ``historical" {\tt DIAForcedSource} records, alert content redundancy could be avoided by only including previously undistributed {\tt DIAForcedSource} data in an alert packet.
+To do this, the Prompt pipeline would have to either compare with the contents of the most recent alert and remove redundant data from the new alert, or flag the {\tt DIAForcedSource} data as ``released" and only add ``unreleased" forced photometry to a alert packet.
+Both of these options would add processing times to the alert generation pipeline, which is undesirable.
 
-Pro: every alert packet contains all the public data and there is no need to provide, e.g., a \emph{publicly accessible} stamps database.
+{\bf Summary --} If the decision is made to remove the historical {\tt DIASource} records from alert packets, the {\tt DIAForcedSource} records of forced photometry should always be added to all alerts.
+This will cause some redundancy in alert contents, but if not included there would be no public access to this scientifically valuable information.
 
-\subsection{Adding Associated {\tt Object} Record Elements}\label{ssec:packets_add}
+\subsubsection{Postage Stamps}\label{sssec:packets_remove_stamps}
 
-The contents of the alert packet are defined in \citeds{LSE-163}.
-They will include {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}, {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds, and {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}.
+Including the postage stamps in the alert packet allows for the instant assessment of the {\tt DIASource} in the difference and template images, which has a variety of scientific applications.
+For example, although real/bogus classification of the difference-image sources will be done prior to alert generation, some brokers may still wish to run additional algorithms on the images -- especially in the case of first detections.
+Brokers might also employ algorithms to classify transients and variables based on the images instead of the photometry (e.g., \citealt{2019PASP..131j8006C}).
+From the postage stamp images, scientifically useful information can be derived about the trails left by moving objects, or from the 2-dimensional luminosity distribution of spatially evolving objects such as comets, which might not be entirely captured by the {\tt DIASource} elements (see Table 1 in \citeds{LSE-163}).
+The images also provide context for the {\tt DIASource}, such as host-galaxy morphology or field crowdedness, that can assist with follow-up observations.
+
+If postage stamps were not created during alert production, the only option would be for users to wait until the images are made available (within 24 hours) in the LSP, which would significantly inhibit alerts-based science goals.
+However, postage stamps are unlikely to be needed for most {\tt DIASources}, the majority of which --- 95\%, according to the breakdown in \citeds{DMTN-102} --- will be stars or asteroids: point sources for which context is not a significant part of their immediate evaluation.
+An efficient compromise might be for the Prompt pipeline to create the postage stamps during alert generation, and then store them in a publicly accessible database (i.e., outside the LSP) where they can be obtained (via {\tt diaSourceId}) with a latency similar to the alert distribution timescale by anyone, from anywhere, at anytime.
+
+The expected maximum download rate from this stamp database is estimated by assuming that all 5 community brokers download $\sim$500 stamp sets per visit: $\sim$2500 stamp sets every $\sim$35 seconds.
+At 18~KB per stamp set, that's a data rate of 10.5 Mbps (for comparison, that's $\sim$5\% of one full alert stream).
+
+{\bf Summary --} The postage stamps could be removed from all alert packets, but they should still be created during Prompt processing and stored in a publicly accessible database where they can be obtained with a latency similar to the alert distribution timescale.
+
+\subsubsection{Pros and Cons of Multiple Alert Formats}\label{sssec:packets_remove_procon}
+
+The above options could either apply to all alert packets, or brokers could choose from the menu and subscribe to a stream with the options they want.
+The latter means that LSST distributes multiple types of alert packet formats, and there are benefits and drawbacks to this ``menu" approach.
+The LSST currently expects to deliver alert streams to five brokers, and there are four alert packet formats on the table: (i) with histories and stamps, (ii) with histories but without stamps, (iii) without histories but with stamps, and (iv) without histories or stamps.
+
+From a science perspective, the benefits to allowing brokers to customize their alert packet is that it potentially enables unique science to be done by brokers (e.g., custom stamp-analysis algorithms), and could enable LSST to support $>$5 brokers which might lead to more alerts-related science in general.
+One drawback might be that in times of heavy load, full-sized packet streams might be more likely to be delayed, which would be unfair to brokers requiring full-size packets, and could disproportionately affect some science goals more than others.
+The fact that not all alert packets are identical might cause some additional bookkeeping considerations for downstream brokers who ingest from multiple community brokers, but should not be detrimental to their science goals.
+
+The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is left for data management.
+
+\subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}
+
+The contents of the alert packet as defined in \citeds{LSE-163} includes the following with respect to associations with {\tt Object} catalog from the most recent data release:
+\begin{itemize}
+\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
+\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
+\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
+\end{itemize}
 For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.
 
-First, this definition of {\tt nearbyObjLnP} seems to only be appropriate for stars.
-For galaxies, the likelihood of a nearby galaxy being the true host is proportional to {\tt nearbyObjDist} divided by the effective radius of the host, or even better, \textcolor{red}{find the definition that uses semi-major and minor axies and the position angle etc.}.
-For this reason, the {\tt nearbyObj} included in the alerts should not simply be the three nearest galaxies -- it should be the \emph{three most likely host galaxies}.
+The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
+For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxy's luminosity profile.
+Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".
 
-Second, the most useful piece of information about a transient's host galaxy is the redshift, because this significantly improves photometric classification algorithms \textcolor{red}{(could cite RAPID or Ashely's work and others)}.
+Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
+This is usually estimated by calculating an offset distance from the nearby galaxies to the transient that is expressed in terms of the galaxy's spatial luminosity profile, and assuming the galaxy with the lowest offset distance is the host.
+The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.
 
-Including the three nearest stars remains appropriate because all stars have the same radial profile.
-For variable stars, \textcolor{red}{the most useful piece of information about its {\tt Object} is probably the mean apparent magnitude?}
+{\bf Effective Radius --} Calculate a separation distance that is the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy (i.e., {\tt kronRad90} in the {\tt Objects} table; \citeds{LSE-163}).
+The nearby galaxy with the lowest separation distance is the most likely host. This will account for the relative sizes of the potential host galaxies, but not their position angles.
+
+{\bf Second Moment --} Calculate a separation distance, as in \citet{2006ApJ...648..868S}, based on the two-dimensional luminosity profile of the galaxy.
+Where $x_{\rm trans},y_{\rm trans}$ is the location of the transient, and $x_{\rm gal},y_{\rm gal}$ is the center of the galaxy (the first moment; {\tt radec} in the {\tt Object} table), the separation distance is $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $x_r = x_{\rm SN} - x_{\rm gal}$ and $y_r = y_{\rm SN} - y_{\rm gal}$.
+The ellipse parameters $C_{xx}$, $C_{yy}$, and $C_{xy}$ can be calculated directly from the second moments of the galaxy luminosity profile ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy} in the {\tt Object} table), e.g., as described in Section 10 of E.
+Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}}.
+The nearby galaxy with the lowest separation distance is the most likely host.
+This option accounts for both the relative sizes and position angles of the potential host galaxies, and requires only a little more processing.
+
+{\bf 2D Algorithms --} There are also other, more complicated methods for identifying the most likely host for a given transient.
+For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
+Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
+The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.
+
+Generally, nearby galaxies with separation distances $>$3, or $>$99\% of the luminosity profile, are not assigned as hosts and the transient is considered "hostless".
+While this cutoff has been appropriate for past samples of $\sim$hundreds of transients, it will not be appropriate for the LSST sample size, and no such cut should be applied when evaluating potential host galaxies for {\tt DIAObjects}.
+
+All of the above host galaxy identification methods can also make use of priors if the transient type is known; for example, the established correlation between core-collapse supernovae and star formation.
+However, such a robust level of host associations are beyond scope for the Prompt pipeline, and the main goal is to provide brokers with sufficient information to assess the potential host association, and decide whether it is useful to the classification and/or prioritization of alerts for follow-up.
+
+{\bf Summary --} For the ten nearest {\tt Object} catalog galaxies, a separation distance should be calculated with respect to the transient location, preferably using the second moments of each galaxy's luminosity profile.
+Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances.
+An analog for the existing element {\tt nearbyObjLnP}, representing the probability, is not necessary.
+This would add {\tt unit64[3]} and {\tt float[3]} to the {\tt DIAObject} catalog and to each alert, but this both tiny and worthwhile.
 
@@
 The application of gzip compression could further reduce the size of a full alert to $\sim$65~KB (80\%; JIRA ticket DM-16280).
-Naively, this seems like it would allow 1 more full stream to be transmitted to a broker, but the time and computational resources required to compress the alert packets has not been considered, \emph{and} this option would then force all brokers to decompress the alerts on arrival which might further delay processing for everyone.
-Science goals requiring low-latency alerts (\S~\ref{sec:latency}) could be negatively impacted.
+Naively, this seems like it would allow 1 more full stream to be transmitted to a broker, and thereby enable more science.
+However, the time and computational resources required to compress the alert packets needs to be considered.
+For example, gzip compression at 50 MB/s to compress $\sim$10000 alerts would take $\sim$10000 $\times$0.08 MB per alert $/$50 MB/s, or $\sim$16 seconds.
+This is significant, considering the alert distribution latency requirement\reqparam{OTT1}\lsrreq{0101}\ossreq{0127}\dmreq{0004} is 60 seconds. 
+Furthermore, compressing the alert packets forces brokers to then decompress the alerts on arrival, which would incur further delay.
+Science goals requiring very low-latency alerts distribution might be negatively impacted by compression -- see \S~\ref{sec:latency} for a deeper discussion of the science drivers for low-latency alerts.
+
+{\bf Summary --} Since packet compression could induce an additional latency, perhaps it should only be done if, e.g., an algorithm could provide a compression rate of $\lesssim$10 MB/s (requiring only a few seconds to compress a visit's worth of alerts), or if the alert distribution latency requirement is relaxed from 1 minute (see \S~\ref{sec:latency}).
+Otherwise, compression would have no impact on alerts-related science goals and so this should be primarily a technically-driven decision.
 
@@
 
-\textcolor{red}{MLG: Can grab all the stuff you put into the CBW paper and put it here instead.}
+One of the strongest -- but also most mysterious -- transients that might significantly benefit from the $60$ second alert timescale are fast radio bursts (FRBs): a millisecond long pulse of coherent emission in the GHz range.
+The emission is dispersed by the inter-galactic medium (IGM), such that the pulse's observed arrival time is frequency dependent.
+Observed FRB dispersion measures of $\rm{DM}\approx100$ to $1000$ $\rm pc\ cm^{-3}$ indicate that their origins are at cosmological distances, with redshifts $z\approx0.1$ to $1$ \citep{2018Natur.562..386S}.
+
+If an optical counterpart is generated by this coherent emission, the time delay between the optical detection and the radio detection is on the timescale of minutes for frequencies $\nu < 500\ {\rm MHz}$, as shown in Figure \ref{fig:sci_frb}.
+This leaves open the possibility for triggering radio follow-up of optical counterpart candidates -- if they exist and are detectable -- and triggering additional follow-up for a serendipitous co-detection of FRB and optical transient.
+A millisecond-long event in the optical would have to be quite bright to be detected by LSST, but studies show that LSST detections are feasible \citep{2016ApJ...824L..18L}.
+\cite{2019ApJ...878...89Y} demonstrate that two theoretical sources for coherent optical emission from FRBs would not be detectable by LSST, but that inverse Compton scattering processes could lead to optical detections with LSST (e.g., from pulsars or masers).
+If FRBs are associated with superluminous supernovae (SLSNe) and young magnetars, then potentially an LSST alert regarding a change in behavior of known SLSNe could be used to trigger radio follow-up.
+However, this case is unlikely to be as sensitive to the LSST alert distribution timescale as the optical emission would be released over a longer time window \citep{2019arXiv191002036L}.
+
+Searches for FRB counterparts in optical images that were serendipitously obtained coincidentally with an FRB detection have only just recently become possible thanks to current wide-field imaging surveys.
+No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.
+Since the rate of FRBs is estimated to be quite high, thousands per day \citep{2016MNRAS.460L..30C}, even after radio detection efficiency is factored in there should be many within the observed LSST volume.
+
+\textcolor{red}{\bf Need an estimate of the number of FRB locations serendipitously observed by LSST within a few minutes window.}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=12cm]{figures/frb_optical_delays.png}
+\caption{The delay time (for coherent emission) between the arrival of the optical and radio photons due to cosmological dispersion, as a function of the frequency of the radio detector.
+The {\it lowest} frequency bands of the Square Kilometer Array (SKA) and the Low-Frequency Array (LOFAR) are marked with vertical dotted and dash-dotted lines, and the 1 minute timescale for LSST alert distribution with a horizontal dashed line.
+The delay time is $\Delta t = k_{\rm DM}\,DM\,(\nu_{\rm low}^{-2} - \nu_{\rm high}^{-2})$, where $k_{DM}=4.149$ $\rm GHz^2\ pc^{-1}\ cm^3\ ms$, ${\rm DM}$ is the dispersion measure in $\rm pc\ cm^{-3}$, $\nu$ is the low and high frequency bands of the observation, and $\Delta t$ is in $\rm ms$.
+We have used the range of FRB dispersion measures as observed by \cite{2018Natur.562..386S}.
+\label{fig:sci_frb}}
+\end{center}
+\end{figure}
 

### 2020-01-07 dmtn-tbd.tex
added to S3, still not done tho

@@
 
-The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is left for data management.
+The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is left for other work.
 
@@
 
-{\bf Generally, science goals that require alerts on timescales shorter than $\sim$5 minutes have that need because they also require follow-up on timescales shorter than $\sim$15--30 minutes.} Most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps).
-Thus, any science goals driving the need for alerts on timescales shorter than $\sim$5 minutes \emph{must also be able to confidently identify their targets with single-epoch single-filter LSST photometry}, or be using additional information (such as host galaxy characteristics, or coincidence with another survey's data).
+{\bf Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.} Most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps).
+Thus, any science goals driving the need for alerts on timescales shorter than $\sim$5 minutes \emph{must also be able to confidently identify their targets with single-epoch single-filter LSST photometry}, or be using additional information (such as host galaxy characteristics or coincidence with another survey's data).
 
 {\bf Target of Opportunity --} It is important to note that for rapid ToO observations with LSST (if they occur), there might not be time to pre-cache slow database queries (e.g., loading slices of the PPDB catalogs) when upcoming pointings are not known in advance, as would happen during regular survey observations.
-This might mean that a 1 minute alert distribution timescale for alerts from ToO imaging surveys cannot be guaranteed, and this should be kept in mind when reading through the science cases for short alert latencies in this section.
-\textcolor{red}{$\leftarrow$ This statement was modified from something ECB mentioned on Slack, double check it's accurate}.
+This might mean that a 1 minute alert distribution timescale for alerts from ToO imaging surveys cannot be guaranteed.
+While this should be kept in mind as a possibility, for the purposes of this study of the \emph{scientific} impacts of the alert latency it assumed to be technically solvable issue.
 
@@
 
-The Transients and Variable Stars (TVS) science collaboration surveyed the science needs of their members with respect to alert latency, and the results are shown in Table \ref{tab:tvs}\footnote{Results courtesy of TVS co-chair Rachel Street.}.
+The Transients and Variable Stars (TVS) science collaboration surveyed the science needs of their members with respect to alert latency, and the results\footnote{Results courtesy of TVS co-chair Rachel Street.} are shown in Table \ref{tab:tvs}.
 The survey asked respondents to choose the maximum tolerable and ideal average delay between the alerts being produced by the LSST data reduction pipeline and the alert information becoming available through the broker service.
@@
 The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
-It is at first confusing that EM-GW was provided as the science driver for such a wide range of alert access timescales, but this is discussed further in \S~\ref{ssec:latency_emgw}, along with several other science cases that rely on rapid access to LSST alerts.
+The fact that these three are also listed as science drivers for longer-latency alert is mainly due to the diversity within the science drivers and the fact that the events have both short- and longer-timescale features.
+Each of these science drivers is discussed in turn below, along with two other science cases that rely on rapid access to LSST alerts: fast radio bursts (FRBs) and solar system objects (SSOs).
 
@@
 During LSST Operations there will already exist a sizable collection of longer-latency follow-up, and it is likely that science will be moving in the direction of pushing to ever earlier detections.
-However, since KNe produce days-long optical afterglows, a 5 minute LSST alerts would likely suffice.
+However, since kilonovae produce days-long optical afterglows, a 5 minute LSST alerts would likely suffice.
 
@@
 Another that predicts emission of a similar color, luminosity, and timescale is the spin-down energy of a long-lived ($10^2$--$10^4$ s) neutron star formed from a binary neutron star merger before its eventual collapse to a black hole \citep{2016ApJ...819...15S}.
-As described in \S~\ref{ssec:latency_grb}, very few events could be detected by serendipitous coincidence by the LSST WFD main survey, and targeted follow-up of well-localized GW events with more appropriate facilities (such as space-based UV/optical imagers) is much more likely to yield detections of this very short lived emission.
+As described in \S~\ref{ssec:latency_grb}, very few events could be detected by serendipitous coincidence by the LSST WFD main survey, and targeted follow-up of well-localized GW events with more appropriate facilities such as space-based UV/optical imagers is much more likely to yield detections of this very short lived emission.
 
@@
 Even now, many imaging follow-up surveys wait for the initial alert (or retraction) to be sent after a round of human vetting of the GW event signal, and this can take several hours.
-With such latencies, the 1 {\it vs.} 5 minute LSST alert timescales becomes inconsequential.
+With such latencies, the question of 1 {\it vs.} 5 minute LSST alert timescales becomes inconsequential.
 
@@
 % (0.836/0.014) * 4/yr = 240/yr
-
 The rate of short GRBs in the local volume ($<$200 Mpc) is estimated to be quite low, $<$4 $\rm yr^{-1}$ \citep{2019arXiv190800100M}.
@@
 
+Photometric data obtained during the first few hours to days of a supernova can constrain the progenitor radius and circumstellar material in the immediate environment, which in turn provides information about the progenitor star, its system, and its final stages of evolution before the explosion.
+For example, \citet{2018Natur.554..497B} model the optical photometric data of the post-shock breakout cooling peak for a massive star's core collapse supernova to constrain the radius and mass of its outer envelope; \citet{2012ApJ...744L..17B} used the first day of photometric observations of a nearby Type Ia (thermonuclear detonation) supernova to show that their progenitor stars must be a compact degenerates; and rare ``blue bumps" in the first few days of a Type Ia supernova's light curve might betray the presence of a non-degenerate companion star \citep[e.g.,][]{2010ApJ...708.1025K,2017ApJ...845L..11H}.
+
+For young supernovae, follow-up observations within a couple of hours are required, and the related science goals are unlikely to be impacted if the alert latency was 5--10 minutes instead of 1 minute.
+
 \subsection{Fast Radio Bursts}\label{ssec:latency_frb}
@@
 The emission is dispersed by the inter-galactic medium (IGM), such that the pulse's observed arrival time is frequency dependent.
-Observed FRB dispersion measures of $\rm{DM}\approx100$ to $1000$ $\rm pc\ cm^{-3}$ indicate that their origins are at cosmological distances, with redshifts $z\approx0.1$ to $1$ \citep{2018Natur.562..386S}.
+Observed FRB dispersion measures of $\rm{DM}\approx100$ to $1000$ $\rm pc\ cm^{-3}$ indicate that they originate at cosmological distances, with redshifts $z\approx0.1$ to $1$ \citep{2018Natur.562..386S}.
 
 If an optical counterpart is generated by this coherent emission, the time delay between the optical detection and the radio detection is on the timescale of minutes for frequencies $\nu < 500\ {\rm MHz}$, as shown in Figure \ref{fig:sci_frb}.
-This leaves open the possibility for triggering radio follow-up of optical counterpart candidates -- if they exist and are detectable -- and triggering additional follow-up for a serendipitous co-detection of FRB and optical transient.
+This leaves open the possibility for triggering radio follow-up of an optical counterpart candidate serendipitously detected by LSST -- if such counterparts exist and are detectable.
 A millisecond-long event in the optical would have to be quite bright to be detected by LSST, but studies show that LSST detections are feasible \citep{2016ApJ...824L..18L}.
 \cite{2019ApJ...878...89Y} demonstrate that two theoretical sources for coherent optical emission from FRBs would not be detectable by LSST, but that inverse Compton scattering processes could lead to optical detections with LSST (e.g., from pulsars or masers).
+Searches for FRB counterparts in optical images that were serendipitously obtained (near-) coincidentally with an FRB detection have only just recently become possible thanks to current wide-field imaging surveys.
+No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.
+Since the rate of FRBs is estimated to be quite high (thousands per day; \citealt{2016MNRAS.460L..30C}) even after radio detection efficiency is factored in there could be many within the observed LSST volume.
+
+For example, if there are 1000 FRB occurring across the full sky every 24 hours, and LSST observes one sixth of the sky during a night's 10 hours of darkness, then there would be $\sim70$ FRB occurring in that area during that time.
+Assuming that both the FRB and the LSST visits are distributed randomly in time and location (although the visits are correlated, obviously), there's a $\sim6\%$ chance that in any given night one visit will be serendipitously coincident with the short lived time-delayed optical counterpart of an FRB.
+If such a source could be recognized as an FRB optical counterpart within a minute, and radio follow-up could begin as soon as possible to detect the radio component and measure the time delay (Figure \ref{fig:sci_frb}).
+However, \textcolor{red}{it is (or rather, MLG is) currently unclear whether (a) there is strong theoretical support for such optical counterparts, and (b) radio telescopes can perform ToO follow-up on $<$15 minute timescales}.
+
 If FRBs are associated with superluminous supernovae (SLSNe) and young magnetars, then potentially an LSST alert regarding a change in behavior of known SLSNe could be used to trigger radio follow-up.
@@
 
-Searches for FRB counterparts in optical images that were serendipitously obtained coincidentally with an FRB detection have only just recently become possible thanks to current wide-field imaging surveys.
-No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.
-Since the rate of FRBs is estimated to be quite high, thousands per day \citep{2016MNRAS.460L..30C}, even after radio detection efficiency is factored in there should be many within the observed LSST volume.
-
-\textcolor{red}{\bf Need an estimate of the number of FRB locations serendipitously observed by LSST within a few minutes window.}
-
 \begin{figure}[h]
 \begin{center}
-\includegraphics[width=12cm]{figures/frb_optical_delays.png}
+\includegraphics[width=15cm]{figures/frb_optical_delays.png}
 \caption{The delay time (for coherent emission) between the arrival of the optical and radio photons due to cosmological dispersion, as a function of the frequency of the radio detector.
@@
 Follow-up on the SS use case for alerts on very short timescales.}
+
+\subsection{Stellar Variables}\label{ssec:latency_stars}
+
+\textcolor{red}{Return to the road map or science book? Surely there must be some stellar use-cases for 1 minute alerts.}
 
@@
 
-Along with the requirement of a 60 second alert distribution timescale (\S~\ref{sec:latency}, the DMS should support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 10,000 alerts per standard visit on average during a given night, and at least 40,000 alerts per single standard visit.
+Along with the requirement of a 60 second alert distribution timescale (\S~\ref{sec:latency}, there are requirements that the DMS should support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 10,000 alerts per standard visit on average during a given night, and at least 40,000 alerts per single standard visit.
 
@@
 These requirements apply to standard visits which should have produced $\leq$40,000 alerts.
-For example, a visit would be considered "delayed" and count towards that 1\% limit if $>$2\% of its alerts were distributed with a latency of $>$60 seconds.
+For example, a visit would be considered ``delayed" and count towards that 1\% limit if $>$2\% of its alerts were distributed with a latency of $>$60 seconds.
 The requirement that no more than 0.1\% of all science visits fail to generate and/or distribute alerts is integrated over all stages of data handling, not just alert distribution, and includes failures at any stage of prompt processing.
@@
 
-For alerts that are not distributed within 60 seconds, how should they be handled?
+For alerts that are not distributed within 60 seconds, how should they be handled, \textbf{from a science perspective?}
 
 It is a requirement that all alerts be stored in an archival database and be available for retrieval\ossreq{0185}.
-The term "available for retrieval" applies to users with data rights and access to the LSST Science Platform.
+The term ``available for retrieval" applies to users with data rights and access to the LSST Science Platform.
 Like all other Prompt data products, the alerts archive will be updated within 24\reqparam{L1PublicT}\lsrreq{0104} hours \citedsp{LSE-29}.
@@
 \section{LSST Alert Filtering Service}\label{sec:lafs}
+
+\textcolor{red}{check if this changed names again to the "simple filtering service"?}
 

### 2020-01-07 dmtn-tbd.tex
drafted S4

@@
 
-Pre-filtered streams might mean the alerts can be distributed to more brokers than numStreams. What kind of pre-filters might suit the brokers?
+This section explores the scientific motivation and impact of offering brokers pre-filtered alert streams. 
+Pre-filtering could reduce the overall bandwidth of alert distribution and potentially allow streams to be transmitted to more brokers, thereby enabling more alerts-related science.
+Providing pre-filtered streams could also lessen the load and resources used by the brokers, enabling more alerts-related science by allowing brokers to spend a smaller fraction of their budget computational resources.
 
- - magnitude or signal-to-noise ratio (affects completeness/purity)
- - region of sky (like one of the LOI for a particular region)
- - static-sky object association
+The types of pre-filters that a broker would use are mainly driven by the science goals of that particular broker; some examples follow.
+A broker might only find it useful to receive alerts for {\tt DIASources} that are likely to be stellar variables, moving objects, or extragalactic transients.
+A broker might only want to receive alerts that could potentially be observed by its associated follow-up resources -- {\tt DIASources} brighter than a given limit or located in a particular region of sky.
+A broker might employ a suite of algorithms that all require, e.g., at least two detections by LSST before any further action is taken, and so would not need any single-detection alerts.
+A broker might be specialized in a science area that is not done with the wide-fast-deep main survey, but with a special program such as the deep drilling fields or target-of-opportunity imaging surveys.
+Finally, a broker might not be interested in alerts obtained during poor weather conditions (e.g., an image quality or sky background violating some limit) -- although, this type of filter might not be entirely science-driven.
 
- Image free streams of alert packets.
- Sources only associated with past alerts.
+For the above reasons, the potential pre-filters that might be scientifically useful -- some of which have been suggested by brokers already -- include:
+\begin{itemize}
+\item {\bf Apparent Magnitude: } only {\tt DIASources} brighter than a specified apparent magnitude (for each of the LSST filters {\it ugrizy}) are transmitted. 
+\item {\bf Region of Sky: } only {\tt DIASources} in a given sky area(s) are transmitted. 
+\item \texttt{\bf DIAObject} {\bf History: } only {\tt DIASources} associated with {\tt DIAObjects} that have at least $N$ previous detections are transmitted, where $N$ can be $\geq1$. 
+\item \texttt{\bf SSObject} {\bf Association: } only {\tt DIASources} that are associated with a {\tt SSObject} are transmitted.
+(Potentially including new, unassociated {\tt DIASources}, because they might also be moving objects).
+\item \texttt{\bf Object} {\bf Association: } only {\tt DIASources} associated with {\tt DIAObjects} that are, in turn, associated with a data release {\tt Object} that meets some specification are transmitted (e.g., associated with a stellar point source).
+\item {\bf Observation Metadata: } only alerts from images obtained with a given set of metadata, e.g., a special observing program or meeting sky condition thresholds, are transmitted.
+\end{itemize}
 
-Provide estimates for how much the alert stream volume would be reduced.
+All of these potential science-driven pre-filters are based on the information contained in the alert itself, and do not impose any additional processing on the Prompt pipeline.
+
+\textcolor{red}{Provide estimates for how much the alert stream volume would be reduced by each of these potential pre-filters using nominal limits.}
 

### 2020-01-08 dmtn-tbd.tex
draft S5 done

@@
 
-\textcolor{red}{Return to the road map or science book? Surely there must be some stellar use-cases for 1 minute alerts.}
+\textcolor{red}{Return to the road map or science book? Surely there must be some stellar use-cases for 1 minute alerts. E.g., microlensing peaks can be real sharp.}
 
 \subsection{Summary}\label{ssec:latency_summary}
+
+There does not, currently, seem to be an unambiguous need for alerts on latency timescales of 1 minute compared to 5 minutes. 
+However, it is not unambiguous that such a need will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events.
 
@@
 
-For the above reasons, the potential pre-filters that might be scientifically useful -- some of which have been suggested by brokers already -- include:
+For the above science drivers, the types of pre-filters that are likely to be scientifically useful -- some of which have been suggested by brokers already -- include:
 \begin{itemize}
-\item {\bf Apparent Magnitude: } only {\tt DIASources} brighter than a specified apparent magnitude (for each of the LSST filters {\it ugrizy}) are transmitted. 
-\item {\bf Region of Sky: } only {\tt DIASources} in a given sky area(s) are transmitted. 
-\item \texttt{\bf DIAObject} {\bf History: } only {\tt DIASources} associated with {\tt DIAObjects} that have at least $N$ previous detections are transmitted, where $N$ can be $\geq1$. 
-\item \texttt{\bf SSObject} {\bf Association: } only {\tt DIASources} that are associated with a {\tt SSObject} are transmitted.
+\item {\bf Apparent Magnitude:} only {\tt DIASources} brighter than a specified apparent magnitude (for each of the LSST filters {\it ugrizy}) are transmitted. 
+\item {\bf Region of Sky:} only {\tt DIASources} in a given sky area (or areas; defined by, e.g., limits in right ascension or declination), are transmitted. 
+\item \texttt{\bf DIAObject} {\bf History:} only {\tt DIASources} associated with {\tt DIAObjects} that have at least $N$ previous detections are transmitted, where $N$ can be $\geq1$. 
+\item \texttt{\bf SSObject} {\bf Association:} only {\tt DIASources} that are associated with a {\tt SSObject} are transmitted.
 (Potentially including new, unassociated {\tt DIASources}, because they might also be moving objects).
-\item \texttt{\bf Object} {\bf Association: } only {\tt DIASources} associated with {\tt DIAObjects} that are, in turn, associated with a data release {\tt Object} that meets some specification are transmitted (e.g., associated with a stellar point source).
-\item {\bf Observation Metadata: } only alerts from images obtained with a given set of metadata, e.g., a special observing program or meeting sky condition thresholds, are transmitted.
+\item \texttt{\bf Object} {\bf Association:} only {\tt DIASources} associated with {\tt DIAObjects} that are, in turn, associated with a data release {\tt Object} that meets some specification are transmitted (e.g., associated with a stellar point source).
+\item {\bf Observation Metadata:} only alerts from images obtained with a given set of metadata, e.g., a special observing program or meeting sky condition thresholds, are transmitted.
 \end{itemize}
@@
 
+% % % % % % % % % % % % % % % % % % % % % % % % % %
+\subsection{LSST Alert Filtering Service}\label{ssec:prefilter_lafs}
+
+As described in \citeds{LSE-163} the LSST will provide an alert filtering service with limited capacity, by which individuals with LSST data rights and access may receive alerts via pre-defined filters and/or create and apply their own filters to the stream \citedsp{LPM-17,LSE-61}.
+It is a requirement that the LSST alert filtering service be able to support\reqparam{numBrokerUsers}\dmreq{0343} at least 100 simultaneous users, and capable of returning 20\reqparam{numBrokerAlerts}\dmreq{0343} full-sized alerts per visit per user.
+User-generated filters may be comprised of, e.g., a series of {\tt if} statements applied to the alert packet elements.
+There are no requirements on the latency of filter execution, which may depend on the complexity of the filter and available resources.
+
+From a science perspective, it would be beneficial to offer any and all of the pre-filtered streams listed above as options for user-generated filters.
+Individual users running small follow-up programs are more likely to have access to smaller-aperture telescopes and thus desire an initial apparent magnitude cut.
+It would save on compute resources to avoid a situation in which, e.g., 50 user-generated brokers all start with an {\tt if} statement to filter the stream to, e.g., $m_r < 20$ mag.
+
+% One type of simple filtering that individual users are more likely to want than brokers is ``watchlist" capability -- lists of individual objects (stars or galaxies) and offset radii that alerts are desired within.
+
 \clearpage
@@
 
-Along with the requirement of a 60 second alert distribution timescale (\S~\ref{sec:latency}, there are requirements that the DMS should support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 10,000 alerts per standard visit on average during a given night, and at least 40,000 alerts per single standard visit.
+The LSST specifications for the DMS require that it support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 40,000 alerts per single standard visit, and that for visits producing $\leq$40,000 alerts \reqparam{sciVisitAlertDelay}\reqparam{sciVisitAlertFailure}\ossreq{0112}\dmreq{0392} no more than 1\% of them fail to have at least 98\% of its alerts distributed within 60 seconds of image readout.
+It is furthermore specified that alert distribution "degrade gracefully" beyond that limit, meaning that visits resulting in an excess of 40,000 of alerts should not cause any DMS downtime \citedsp{LSE-30,LSE-61}.
 
-It is also a requirement\reqparam{sciVisitAlertDelay}\reqparam{sciVisitAlertFailure}\ossreq{0112}\dmreq{0392} that no more than 1\% of all standard visits fail to have at least 98\% of its alerts distributed within 60 seconds of image readout, and that no more than 0.1\% of all standard visits fail to distribute alerts.
+This leaves the open question of what, from a science perspective, is the optimal way of dealing with delayed alerts that also meets the DMS specification of a ``graceful degradation"?
+We leave the aspects of technical implementation of a ``graceful degradation", such as distributing delayed alerts and alert archive storage access, for elsewhere and here just consider the science implications for a delayed timescale.
 
-These requirements apply to standard visits which should have produced $\leq$40,000 alerts.
-For example, a visit would be considered ``delayed" and count towards that 1\% limit if $>$2\% of its alerts were distributed with a latency of $>$60 seconds.
-The requirement that no more than 0.1\% of all science visits fail to generate and/or distribute alerts is integrated over all stages of data handling, not just alert distribution, and includes failures at any stage of prompt processing.
+{\bf Next-opportunity distribution via the alert stream.} As discussed in Section \ref{sec:latency}, there remains plenty of science goals that do not absolutely require alert distribution in 1 minute.
+Distributing delayed alerts via the stream would still enable plenty of science.
+The brokers might prefer to have delayed alerts clearly flagged to properly process them (e.g., some filtering and processing done by brokers might only be appropriate for alerts delivered within a given latency).
 
-It is furthermore specified that alert distribution "degrade gracefully" beyond that limit, meaning that visits resulting in an excess of alerts should not cause any DMS downtime \citedsp{LSE-30,LSE-61}.
+{\bf Next-morning distribution via the alert stream.} This refers to an option to collect all delayed alerts during the night and then releasing them (perhaps on a new ``topic") in the morning after survey operations have ended for the night.
+From a science perspective this is not as useful as next-opportunity distribution, but if it is preferred for technical reasons it would enable more science than the option below.
 
-In crowded fields there might be $>$40,000 difference-image sources.
+{\bf Do not distribute delayed alerts; send directly to archive.} There is no scientific merit in not distributing delayed alerts, and two further drawbacks: the alert archive update timescale is 24 hours\reqparam{L1PublicT}, significantly slower than next-morning distribution, and the alert archive might only be accessible by brokers with data rights since it is part of the LSST Data Facility.
 
-For alerts that are not distributed within 60 seconds, how should they be handled, \textbf{from a science perspective?}
-
-It is a requirement that all alerts be stored in an archival database and be available for retrieval\ossreq{0185}.
-The term ``available for retrieval" applies to users with data rights and access to the LSST Science Platform.
-Like all other Prompt data products, the alerts archive will be updated within 24\reqparam{L1PublicT}\lsrreq{0104} hours \citedsp{LSE-29}.
-
-\textcolor{red}{From the community post brokers FAQ $\rightarrow$}The retention period in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.
-Brokers with data rights can retrieve old alert packets from the Alerts Database.
-However, note that the requirements on user access to the Alert Database are not yet fully developed, and may be limited in terms of how alert packets may be queried (e.g., alert ID), or by the bulk download capabilities (DMTN-102 1).
-
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{LSST Alert Filtering Service}\label{sec:lafs}
-
-\textcolor{red}{check if this changed names again to the "simple filtering service"?}
-
-It is a requirement that the LSST alert filtering service be able to support\reqparam{numBrokerUsers}\dmreq{0343} at least 100 simultaneous users.
-The LAFS will be capable of returning 20\reqparam{numBrokerAlerts}\dmreq{0343} full-sized alerts per visit per user.
-
-The LSST alerts filtering service is a mechanism by which users --- individuals with LSST data rights and access --- can receive alerts via pre-defined filters that have been optimized for established transient classifications such as supernovae and/or create and apply their own filters to the stream \citedsp{LPM-17,LSE-61}.
-
-Which options might be appropriate for implementation with the LAFS?
-A few of these options might also be available to users of the LSST alert filtering service (\S~\ref{sec:LAFS}).
+% MLG: commented out technical implementation aspects.
+%The retention period of alert packets in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.
+% It is a requirement that all alerts be stored in an archival database and be available for retrieval\ossreq{0185}.
+%The term ``available for retrieval" applies to users with data rights and access to the LSST Science Platform.
+%Like all other Prompt data products, the alerts archive will be updated within 24\reqparam{L1PublicT}\lsrreq{0104} hours \citedsp{LSE-29}.
+% However, note that the requirements on user access to the alerts archive are not yet fully developed, and may be limited in terms of how alert packets may be queried (e.g., alert ID), or by the bulk download capabilities.
 

### 2020-01-08 dmtn-tbd.tex
new S1

@@
 
+% Cite requirements using these macros.
+% \lsrreq \ossreq \dmreq \reqparam
+
 % Create the title page.
@@
 
-The LSST Science Requirements Document \citedsp{LPM-17} specifies that information about the detections of transient, variable, and moving objects be released promptly as a data stream.
+% The LSST Science Requirements Document \citedsp{LPM-17} specifies that information about the detections of transient, variable, and moving objects be released promptly as a data stream.
 
-The purpose of this document is to consider the science motivation for, and impact of, the various options for alert packet contents and distribution timescales.
-
-The contents and delivery of alert packets play a major role in the scientific impact potential of LSST, in part because they are the only LSST data product that is both world public --- they are exempt from any proprietary period --- and publicly accessible -- they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
+The purpose of this document is to consider the science motivation for, and impact of, various options for alert packet contents and their distribution timescale.
+The contents and delivery of alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public, as they are exempt from any proprietary period, and publicly accessible because they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
 In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
@@
 This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
-The size and latency of alerts could therefore affect the ultimate number of supported community alert brokers.
+Reducing the packet size or increasing (slightly) the latency could enable the stream to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
+The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
+Reducing the contents should only be done if that information is easily and quickly obtainable elsewhere.
 
-This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
-\begin{itemize}
-\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
-\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
-\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
-\item DM-20298, Revise the contents of the alert packets (Graham)
-\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
-\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
-\end{itemize}
+{\bf Document Overview ---} \S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
+\S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
+\S~\ref{sec:prefilter} lists the expected types of pre-filtered streams that could improve brokers' scientific productivity, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
+%In \S~\ref{sec:packets} we consider the potential science impact of options for the alert contents that could reduce packet size, and a few small additions that would assist with source classification.
+In \S~\ref{sec:latency} we consider the various science drivers for an alert distribution latency of 1 minute, and the potential scientific impact of a relaxed (e.g., 5 minute) latency.
+In \S~\ref{sec:prefilter} and \ref{sec:graceful} we consider the potential effects on scientific productivity regarding options to provide brokers with pre-filtered streams and delayed alert packets.
 
-% \lsrreq \ossreq \dmreq \reqparam
+%This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
+%\begin{itemize}
+%\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
+%\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
+%\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
+%\item DM-20298, Revise the contents of the alert packets (Graham)
+%\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
+%\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
+%\end{itemize}
 

### 2020-01-09 dmtn-tbd.tex
first draft ready for input

@@
 \author{%
-M.~L.~Graham, T. Pritchard, and the Data Management System Science Team
+M.~L.~Graham and the Data Management System Science Team, with contributions from the Transients and Variable Stars Science Collaboration
 }
@@
   \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
+  \addtohist{0.1}{2020-01-09}{First draft ready for input from others.}{Melissa Graham}
 }
@@
 \S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
-\S~\ref{sec:prefilter} lists the expected types of pre-filtered streams that could improve brokers' scientific productivity, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
-%In \S~\ref{sec:packets} we consider the potential science impact of options for the alert contents that could reduce packet size, and a few small additions that would assist with source classification.
-In \S~\ref{sec:latency} we consider the various science drivers for an alert distribution latency of 1 minute, and the potential scientific impact of a relaxed (e.g., 5 minute) latency.
-In \S~\ref{sec:prefilter} and \ref{sec:graceful} we consider the potential effects on scientific productivity regarding options to provide brokers with pre-filtered streams and delayed alert packets.
+\S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
 
@@
 
-The section explores the science impacts of various options for alert packet contents, such as the $\sim$12 month history of {\tt DIASource} records, the image stamps, and the static-sky {\tt Object} catalog associations.
-Reducing the size of an alert packet might be scientifically beneficial if it enables the stream to be transmitted to more community brokers, thereby leading to more science being done with alerts.
-It might also be beneficial if it enables brokers to spend less time developing software or spend less of their computational resources to remove parts of the alert packet they do not need prior to processing, and spend more on developing and running algorithms for various alerts-related science goals.
-\S~\ref{ssec:packets_remove} and \ref{ssec:packets_compress} consider reducing the size of alert packets by relocating some data or compressing the packets, respectively. 
-A small increase to the size of the alert packet is not necessarily going to be scientifically detrimental if the information added enables science, and \S~\ref{ssec:packets_add} considers some small additions regarding the association of {\tt DIASources} with the data release {\tt Object} catalog.
+In this section the scientific impacts of various options to reduce the size of alert packets (in order to, e.g., enable wider alert distribution, facilitate broker processing) are considered, as are potential additions to the alert packet contents which would improve the user's ability to classify and prioritize alert sources for follow-up.
+The main conclusions of the following discussion are:
+\begin{itemize}
+\item {\bf Removing Histories (\S~\ref{sssec:packets_remove_hist}) ---} Removing the {\tt DIASource} histories from the alert packets that are transmitted to brokers that request this modification could be scientifically beneficial, but a list of all {\tt DIASource} identifiers associated with the {\tt DIAObject} at the time of alert generation must be added to those packets.
+\item {\bf Removing Stamps (\S~\ref{sssec:packets_remove_stamps}) ---} Removing the image stamps from all alert packets could be scientifically beneficial, but stamps should still be made during alert production and made publicly retrievable with a latency similar to that of alert distribution.
+\item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
+\item {\bf Adding Object Association Information (\S~\ref{ssec:packets_add}) ---} To better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
+\item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
+\end{itemize}
 
@@
 Removing the {\tt DIASource} record history would completely inhibit this kind of analysis for brokers that do not have such storage or processing capabilities.
+
 There is also the issue that the association of {\tt DIASources} into {\tt DIAObjects} may change over time, for some objects.
@@
 {\bf Summary --} The {\tt DIASource} record histories should not be removed from all alert packets, but could be optionally removed at the request of a broker.
-However, if the record histories are removed, a list of all {\tt DIASource} identifiers (the {\tt diaSourceId}, as in Section 3.3.1 of \citeds{LSE-163}) which are currently associated with the {\tt DIAObject} to which the alert-spawning {\tt DIASource} is associated should be provided (note that this list is \emph{not} part of the {\tt DIAObject} record).
+However, if the record histories are removed, a list of all {\tt DIASource} identifiers (the {\tt diaSourceId}, as in Section 3.3.1 of \citeds{LSE-163}) which are currently associated with the {\tt DIAObject} to which the alert-spawning {\tt DIASource} is associated should be provided (note that this list is \emph{not} currently part of the {\tt DIAObject} record).
 
@@
 However, it is not always redundant information that was previously released in an alert, like the {\tt DIASource} record history.
+
 There are two main cases where forced photometry is performed, as described in \citeds{LSE-163}.
@@
 
-The above options could either apply to all alert packets, or brokers could choose from the menu and subscribe to a stream with the options they want.
-The latter means that LSST distributes multiple types of alert packet formats, and there are benefits and drawbacks to this ``menu" approach.
-The LSST currently expects to deliver alert streams to five brokers, and there are four alert packet formats on the table: (i) with histories and stamps, (ii) with histories but without stamps, (iii) without histories but with stamps, and (iv) without histories or stamps.
+The histories and/or stamps could be removed from all alert packets, or only at the request of a broker.
+In the latter case, LSST might end up distributing four types of alert packet formats to its five brokers: (i) with histories and stamps, (ii) with histories but without stamps, (iii) without histories but with stamps, and (iv) without histories or stamps.
 
-From a science perspective, the benefits to allowing brokers to customize their alert packet is that it potentially enables unique science to be done by brokers (e.g., custom stamp-analysis algorithms), and could enable LSST to support $>$5 brokers which might lead to more alerts-related science in general.
-One drawback might be that in times of heavy load, full-sized packet streams might be more likely to be delayed, which would be unfair to brokers requiring full-size packets, and could disproportionately affect some science goals more than others.
-The fact that not all alert packets are identical might cause some additional bookkeeping considerations for downstream brokers who ingest from multiple community brokers, but should not be detrimental to their science goals.
+From a science perspective, the benefits to allowing brokers to customize their alert packet is that it potentially enables unique science to be done by brokers (e.g., very rapid custom stamp-analysis algorithms), and could enable LSST to support $>$5 brokers which might lead to more alerts-related science in general.
 
-The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is left for other work.
+One drawback might arise during times of heavy load, such as many visits with $>$40,000 alerts (\S~\ref{sec:graceful}), \emph{if}, and only if, full-sized packet streams are more likely to be delayed.
+Whether this becomes a possibility will depend on the technical aspects of how the multiple types of alerts are generated and transmitted, and also on the final decisions about latency (\S~\ref{sec:latency}) and delayed distribution (\S~\ref{sec:graceful}).
+This may well never come to pass, but if it did, it could be unfair to brokers requiring full-size packets.
+While some brokers are being designed for a wide variety of science goals, others are more specifically tailored.
+If it is the specifically-tailored broker that requires full-sized alerts, then experiencing delayed transmission more frequently than other brokers could put some science goals at a disadvantage. 
+Furthermore, the fact that not all alert packets are identical might cause some additional bookkeeping considerations for downstream brokers who ingest from multiple community brokers, but that should not be detrimental to their science goals.
+The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is beyond the scope of this document, in which we focus on science impacts.
+
+{\bf Summary --} Offering multiple alert formats to brokers is scientifically beneficial provided that full-sized alerts are not predisposed to delay.
 
@@
 
+This section considers the scientific motivation for the 1 minute alert distribution latency, and the impact to alerts-related science if that latency was extended to, e.g., 5 minutes.
 It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
+However, if it is possible to relax this requirement without a negative impact on science then it might be possible for LSST to transmit alerts to $>$5 brokers (and/or repurpose human and computational resources dedicated to the 1 minute latency), which in turn might enable more alerts-related science.
 
-If it is possible to relax this requirement to, e.g., 5 minutes -- \emph{without risk to the science goals} -- then it might be possible to support more community brokers and/or repurpose the human and computational resources that would otherwise be dedicated to delivering 1 minute alert distribution.
-This section attempts to gather the science goals which might potentially require 1 minute alert distribution and could not be satisfied with a longer latency.
+Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
+Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
+For the purposes of this discussion, an optimistic outlook is assumed, and the difficulty in identifying targets with a single epoch of LSST photometry is not considered as a reason to downgrade the importance of a 1 minute alert latency.
 
-{\bf Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.} Most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps).
-Thus, any science goals driving the need for alerts on timescales shorter than $\sim$5 minutes \emph{must also be able to confidently identify their targets with single-epoch single-filter LSST photometry}, or be using additional information (such as host galaxy characteristics or coincidence with another survey's data).
-
-{\bf Target of Opportunity --} It is important to note that for rapid ToO observations with LSST (if they occur), there might not be time to pre-cache slow database queries (e.g., loading slices of the PPDB catalogs) when upcoming pointings are not known in advance, as would happen during regular survey observations.
-This might mean that a 1 minute alert distribution timescale for alerts from ToO imaging surveys cannot be guaranteed.
-While this should be kept in mind as a possibility, for the purposes of this study of the \emph{scientific} impacts of the alert latency it assumed to be technically solvable issue.
+After a consideration of the following science cases that are most likely to require rapid follow-up (summary below), it appears that most of them would be satisfied without risk if the latency was increased from 1 to 5 minutes.
+However, it is absolutely not unambiguously clear that a need for 1 minute alerts will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events over the next couple of years.
+\begin{itemize}
+\item {\bf A TVS Survey (\S~\ref{ssec:latency_tvs})} found that only $<$10\% of respondents reported that requiring alerts within 1 minute was necessary for their science goals.
+\item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger).
+However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
+\item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.
+\item {\bf Fast Radio Bursts (\S~\ref{ssec:latency_frb}) ---} \textcolor{red}{Tentatively,} some models predict that a short, bright burst of optical emission from the physical mechanism powering FRBs could be detected by LSST 5 to 15 minutes prior to the radio signal.
+If such optical events occur with significant frequency, are bright and consistently detectable, and can be identified with a single-filter detection, and if radio telescopes can respond within minutes, then LSST is the most promising future facility for this science, which requires a 1 minute alert distribution latency.
+\item {\bf Young Supernovae (\S~\ref{ssec:latency_ysne}) ---} The science goals require follow-up in an hour or longer, and a 5 minute latency would not have a negatively impact.
+\item {\bf Solar System Objects (\S~\ref{ssec:latency_sso}) ---} \textcolor{red}{TBD.}
+\item {\bf Stellar Variables (\S~\ref{ssec:latency_stars}) ---} \textcolor{red}{TBD.}
+\end{itemize}
 
 \subsection{TVS Survey Results Regarding Alerts Timescales}\label{ssec:latency_tvs}
+{\it Contents contain the results of a TVS survey administered by Rachel Street.}
 
@@
 Note that the pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.
-
-Only 20\% (10\%) of respondents report that $<$10 minutes ($\leq$1 minute) is an ideal average delay, and whereas 70\% report that $>$30 minutes would be a sufficient average latency.
-The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
-The fact that these three are also listed as science drivers for longer-latency alert is mainly due to the diversity within the science drivers and the fact that the events have both short- and longer-timescale features.
-Each of these science drivers is discussed in turn below, along with two other science cases that rely on rapid access to LSST alerts: fast radio bursts (FRBs) and solar system objects (SSOs).
 
@@
 
+Only 20\% (10\%) of respondents report that $<$10 minutes ($\leq$1 minute) is an ideal average delay, and whereas 70\% report that $>$30 minutes would be a sufficient average latency.
+The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
+The fact that these three are also listed as science drivers for longer-latency alert is mainly due to the diversity within the science drivers and the fact that the events have both short- and longer-timescale features.
+Each of these science drivers is discussed in turn below, along with two other science cases that rely on rapid access to LSST alerts: fast radio bursts (FRBs) and solar system objects (SSOs).
+
 \subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
-% From Slack conversation, Federica Bianco, Om Sharan Salafia, Eric Bellm
+{\it Some contents revised from a Slack conversation in the TVS chanel between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}
 
@@
 There are two additional issues related to ToO for EMGW events which might make 1 minute alerts necessary or impossible.
-First, as described above, for a ToO imaging survey with LSST there might not be time to pre-load catalogs for the targeted fields.
+
+First, for rapid ToO observations when visit coordinates are not known in advance, there might not be time to pre-cache the slow database queries that are needed during alert production processing (e.g., loading slices of the PPDB catalogs).
+This might mean that an alert distribution latency cannot be guaranteed for ToO imaging surveys. 
+For the purposes of this science-driven discussion, assume this to be a surmountable technical implementation issue.
+
 The second is that, at least for run O3, the GW event detection system itself issues preliminary alerts within 1--10 minutes, and these preliminary alerts are often retracted and do not always have the sky localization.
@@
 
-For short gamma-ray bursts (sGRBs) which are thought to be the mergers of two neutron stars, or a neutron star and a black hole, as the jet propagates through the ejecta material a hot cocoon is formed.
-When cocoon and jet break out, along with sGRB can can be observed as a blue transient with an absolute peak brightness of -12 to -15 magnitudes that lasts for $10^3$--$10^4$ seconds (e.g., \citealt{2018MNRAS.473..576G}).
+{\it \textcolor{red}{MLG: no outside input on this yet, is needed to be confirmed by a GRB person.} }
+
+Short gamma-ray bursts (sGRBs) are thought to be the mergers of two neutron stars, or a neutron star and a black hole.
+As the jet propagates through the ejecta material a hot cocoon is formed.
+When cocoon and jet break out, along with the sGRB can be observed a blue optical transient with an absolute peak brightness of -12 to -15 magnitudes which lasts for $10^3$--$10^4$ seconds (e.g., \citealt{2018MNRAS.473..576G}).
 For an LSST detection limit of $r\sim24$ mag, a detection limit of $r\lesssim-12$ ($r\lesssim-15$) mag corresponds to distances of $\lesssim160$ ($\lesssim630$) Mpc, or redshifts $z\lesssim0.035$ ($z\lesssim0.14$).
-Observing the diversity of cocoon emission will require multi-band follow-up within $10^3$ seconds, or $15$ minutes, and in this particular case a 5 minute alert distribution latency is too long to include the shorter events (or the fainter events for which only the peak would be visible).
+Observing the diversity of cocoon emission for a sample of sGRBs would require regular and consistent multi-band follow-up within $10^3$ seconds, or $15$ minutes.
+In such cases where a sGRB is observed in a time/area that is serendipitously coincident LSST visit, science would be maximized if the alert confirming the presence of an optical counterpart was distributed to brokers within 1 minute.
+However, what is the expected frequency of this situation occurring during LSST Operations?
 
@@
 The rate of short GRBs in the local volume ($<$200 Mpc) is estimated to be quite low, $<$4 $\rm yr^{-1}$ \citep{2019arXiv190800100M}.
-Scaling this up to the co-moving volume within 630 Mpc is $\sim$240 $\rm yr^{-1}$, all-sky.
-In the baseline main survey, about a sixth of the sky can be observed in a given night, which is $\lesssim$0.1 event in the WFD main survey area per night.
-With $\sim$1000 visits per night, and $\sim$10 visits every 5 minutes, there's a $10^{-3}$ chance that the location of a sGRB will be serendipitously observed within 5 minutes of the event on a given night.
-So this could happen once every 3 years, or $\sim$3 times over the 10-year LSST main survey.
+Scaled up to the co-moving volume within 630 Mpc, this rate is $\sim$240 $\rm yr^{-1}$, all-sky.
+In the baseline main survey about a sixth of the sky can be observed in a given night, which implies $\sim$1 sGRBs with detectable optical counterparts in the WFD survey per night for every $\sim$10 nights of observations.
+The probability that one of the 1000 visits per night will occur during the 5 minutes in which an alert would be useful for triggering rapid follow-up is $<10^{-2}$.
+Very roughly, we might expect this serendipitous coincidence to occur twice a year at most, or $\sim$20 times over the 10-year LSST main survey.
 
-For this particular science case, LSST seems to be the wrong tool for the job.
+% 10 visits per 5 minute block, or 100 "blocks" per night
+% 100x100 potential sky area/time options for the 100 blocks, x10 days
+% of the options, 100 blocks per night x 10 nights would be done
+% and then randomly draw just the 1 sGRB
+% scipy.stats.hypergeom(100000,1000,1)
+% 0 0.9900 
+% 1 0.0100 
+% 2 0.0000
+
+For discovering the optical counterparts of sGRBs, the LSST wide-fast-deep survey is simply not the most efficient future facility for the job.
 As described by \citet{2018MNRAS.473..576G}, a rapid search at the location of sGRBs with a UV satellite such as ULTRASAT would be ideal.
@@
 
-\subsection{Young Supernovae}\label{ssec:latency_ysne}
-
-Photometric data obtained during the first few hours to days of a supernova can constrain the progenitor radius and circumstellar material in the immediate environment, which in turn provides information about the progenitor star, its system, and its final stages of evolution before the explosion.
-For example, \citet{2018Natur.554..497B} model the optical photometric data of the post-shock breakout cooling peak for a massive star's core collapse supernova to constrain the radius and mass of its outer envelope; \citet{2012ApJ...744L..17B} used the first day of photometric observations of a nearby Type Ia (thermonuclear detonation) supernova to show that their progenitor stars must be a compact degenerates; and rare ``blue bumps" in the first few days of a Type Ia supernova's light curve might betray the presence of a non-degenerate companion star \citep[e.g.,][]{2010ApJ...708.1025K,2017ApJ...845L..11H}.
-
-For young supernovae, follow-up observations within a couple of hours are required, and the related science goals are unlikely to be impacted if the alert latency was 5--10 minutes instead of 1 minute.
-
 \subsection{Fast Radio Bursts}\label{ssec:latency_frb}
 
-One of the strongest -- but also most mysterious -- transients that might significantly benefit from the $60$ second alert timescale are fast radio bursts (FRBs): a millisecond long pulse of coherent emission in the GHz range.
+{\it Some contents based on text from and discussions with Tyler Pritchard (TVS member).}
+
+One of the most likely -- but also most mysterious -- transients that might significantly benefit from the $60$ second alert timescale are fast radio bursts (FRBs): a millisecond long pulse of coherent emission in the GHz range.
 The emission is dispersed by the inter-galactic medium (IGM), such that the pulse's observed arrival time is frequency dependent.
@@
 No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.
+
 Since the rate of FRBs is estimated to be quite high (thousands per day; \citealt{2016MNRAS.460L..30C}) even after radio detection efficiency is factored in there could be many within the observed LSST volume.
-
 For example, if there are 1000 FRB occurring across the full sky every 24 hours, and LSST observes one sixth of the sky during a night's 10 hours of darkness, then there would be $\sim70$ FRB occurring in that area during that time.
-Assuming that both the FRB and the LSST visits are distributed randomly in time and location (although the visits are correlated, obviously), there's a $\sim6\%$ chance that in any given night one visit will be serendipitously coincident with the short lived time-delayed optical counterpart of an FRB.
+Assuming that both the FRB and the LSST visits are distributed randomly in time and location (although the visits are correlated, obviously), there's a $\sim6\%$ chance that in any given night one visit will be serendipitously coincident with the expected arrival of a short lived optical precursor to the FRB.
 If such a source could be recognized as an FRB optical counterpart within a minute, and radio follow-up could begin as soon as possible to detect the radio component and measure the time delay (Figure \ref{fig:sci_frb}).
@@
 
+\subsection{Young Supernovae}\label{ssec:latency_ysne}
+
+Photometric data obtained during the first few hours to days of a supernova can constrain the progenitor radius and circumstellar material in the immediate environment, which in turn provides information about the progenitor star, its system, and its final stages of evolution before the explosion.
+For example, \citet{2018Natur.554..497B} model the optical photometric data of the post-shock breakout cooling peak for a massive star's core collapse supernova to constrain the radius and mass of its outer envelope; \citet{2012ApJ...744L..17B} used the first day of photometric observations of a nearby Type Ia (thermonuclear detonation) supernova to show that their progenitor stars must be a compact degenerates; and rare ``blue bumps" in the first few days of a Type Ia supernova's light curve might betray the presence of a non-degenerate companion star \citep[e.g.,][]{2010ApJ...708.1025K,2017ApJ...845L..11H}.
+
+For young supernovae, follow-up observations within a couple of hours are required, and the related science goals are unlikely to be impacted if the alert latency was 5--10 minutes instead of 1 minute.
+
 \subsection{Solar System Objects}\label{ssec:latency_sso}
@@
 
-\subsection{Summary}\label{ssec:latency_summary}
-
-There does not, currently, seem to be an unambiguous need for alerts on latency timescales of 1 minute compared to 5 minutes. 
-However, it is not unambiguous that such a need will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events.
-
 \clearpage
@@
 
-The types of pre-filters that a broker would use are mainly driven by the science goals of that particular broker; some examples follow.
-A broker might only find it useful to receive alerts for {\tt DIASources} that are likely to be stellar variables, moving objects, or extragalactic transients.
-A broker might only want to receive alerts that could potentially be observed by its associated follow-up resources -- {\tt DIASources} brighter than a given limit or located in a particular region of sky.
-A broker might employ a suite of algorithms that all require, e.g., at least two detections by LSST before any further action is taken, and so would not need any single-detection alerts.
-A broker might be specialized in a science area that is not done with the wide-fast-deep main survey, but with a special program such as the deep drilling fields or target-of-opportunity imaging surveys.
-Finally, a broker might not be interested in alerts obtained during poor weather conditions (e.g., an image quality or sky background violating some limit) -- although, this type of filter might not be entirely science-driven.
+Depending on the science goals of a given broker and/or their users, a broker might find it useful to only receive alerts for {\tt DIASources} which:
+\begin{itemize}
+\item are likely to have a certain physical origin, e.g., stellar variable, moving objects
+\item could potentially be observed by the follow-up resources its users have access to
+\item have at least one prior detection by LSST (because, e.g., their algorithms all require photometric confirmation or a change in brightness)
+\item originate in images obtained as part of a special program such as the deep drilling fields or target-of-opportunity imaging surveys
+\item were not obtained during poor weather conditions (e.g., an image quality or sky background violating some limit)
+\end{itemize}
 
@@
 All of these potential science-driven pre-filters are based on the information contained in the alert itself, and do not impose any additional processing on the Prompt pipeline.
-
-\textcolor{red}{Provide estimates for how much the alert stream volume would be reduced by each of these potential pre-filters using nominal limits.}
+They all also have the potential to considerably reduce the number of transmitted alerts.
+Due to the exponential relationship between the number of variable stars and their variability amplitude, and also that of volume and distance modulus, an apparent magnitude limit $\sim1$ mag brighter than the nominal $5{\sigma}$ {\tt DIASource} detection limit could reduce the alert stream data rate by $\gtrsim50$\%.
+Limiting to alerts that are associated with past detections or certain types of objects could similarly reduce the alert stream data rate.
+These types of pre-filers would reduce the number of alerts per visit for all visits -- the average bandwidth -- which might enable alert distribution to extend to more brokers and thus enable more alerts-related LSST science.
+Pre-filters based on their sky location or observational metadata would probably either deliver all or none of the alerts for a given visit, causing fluctuations in the bandwidth but not necessarily in a way that would allow for additional streams to be transmitted.
 

### 2020-01-10 dmtn-tbd.tex
statments re. stamp size

@@
 {\bf Summary --} The postage stamps could be removed from all alert packets, but they should still be created during Prompt processing and stored in a publicly accessible database where they can be obtained with a latency similar to the alert distribution timescale.
+
+\textcolor{red}{{\bf Could Postage Stamps Be Smaller? --}}
+\citeds{LSE-163} describes the size of the postage stamp to be at least $6\arcsec \times 6\arcsec$ ($30\times30$ pixels), or large enough to encompass the entire footprint of the {\tt DIASource} (e.g., if trailed, extended).
+That is large enough to contain the angular diameter (30 kpc) of a Milky Way-like galaxy at a redshift of $z\approx0.5$, which will be a fairly typical supernova host galaxy in the LSST data set.
+Since context is a science driver for producing stamps, it would not be desirable for them to be much smaller than $6\arcsec \times 6\arcsec$.
 

### 2020-01-10 dmtn-tbd.tex
added a note re footprints

@@
 Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
+\textcolor{red}{\citeds{LSE-163} does mention that footprints will be generated for {\tt Objects}, but it doesn't look like they'd be stored in the {\tt Object} table; find out if these footprints would be appropriate/available to use here.}
 The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.

### 2020-01-10 dmtn-tbd.tex
small ammendments

@@
 \subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
-{\it Some contents revised from a Slack conversation in the TVS chanel between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}
+{\it Some contents revised from a Slack conversation between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}
 
@@
 
-{\it Some contents based on text from and discussions with Tyler Pritchard (TVS member).}
+{\it Some contents based on text from and discussions with Tyler Pritchard.}
 

### 2020-02-27 draft-alertcontents.tex
renamed files

--- 2020-01-10 dmtn-tbd.tex
@@
 
+\textcolor{red}{MLG notes from DESC: not all alerts need them so it could be, e.g., for new DIASources only, non-stellar DIASources, etc.
+with the rest going to a server that does not have the 1 minute latency requirements.}
+
 Including the postage stamps in the alert packet allows for the instant assessment of the {\tt DIASource} in the difference and template images, which has a variety of scientific applications.
@@
 \subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}
+
+\textcolor{red}{MLG notes from DESC: \\ (1) Check "Host Galaxy Identification for Supernova Surveys" \citet{2016AJ....152..154G}.
+\\ (2) Alex Gagliano is working on host association as a part of classification that does a PCA on galaxy properties, and found that color and morphology are the first two PCA components, and that the hosts of SNIa SNII etc.
+occupy different parts of the PC2 vs PC1 plane.
+His paper is in prep still.
+\\ (3) As MWV was trying to tell you, host association depends on the redshift of the object (i.e., you want to do it in absolute distance); obviously that won't be in the alert so we can't make it part of the "3 nearest", but you'll probably want to mention that specifically.}.
 
@@
 \subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
-{\it Some contents revised from a Slack conversation between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}
+{\it Some contents revised from a Slack conversation between Federica Bianco, Om Sharan Salafia, and Eric Bellm.} \textcolor{red}{MLG notes from DESC: the case for optical counterparts to neutrino events falls in this category as well, with perhaps the same challenge of them being not instantaneous so there is already some $>10$ minute delay such that 60s optical alerts from any follow-up aren't the bottleneck.}
 
@@
 {\it \textcolor{red}{MLG: no outside input on this yet, is needed to be confirmed by a GRB person.} }
+
+\citet{2018ApJ...854L..13H} searched for fast-fading transients in archival data from the intermediate Palomar Transient Factory (iPTF) and found 50 candidates, three of which were confirmed to be the afterglows of long GRBs.
+Of these three, two were in the archival data because the iPTF had done dedicated follow-up observations of Fermi detections, and one was serendipitous.
+They estimate an all-sky rate of such optical transients (peak $m<18$ mag, fade by at least 2 mag in 3 hours) of $680^{+2236}_{-119}$ events per year (similar to that of long GRBs).
+\textcolor{red}{MLG: still need some LSST projection here.}
 
@@
 
+\citet{2006ApJ...644L..63K}, the discovery of three $<$hour optical transients in the Deep Lens Survey (one confirmed M star and two with potentially extragalactic origin), has been invoked as motivation for the 30 second exposure times and 60 second alert production latency -- e.g., the Deep Lens Survey is referenced without citation in the LSST Science Requirements Document.
+More recently, \citet{2013ApJ...779...18B} searched for fast optical transients in the Pan-STARRS1 Medium-Deep Survey identified 19 fast transients: 11 M dwarf flares and 8 likely solor system objects.
+\citet{2018ApJ...854L..13H} searched for fast-fading transients in archival data from the Palomar Transient Factor and found 50 candidates: three GRB afterglows (\S~\ref{ssec:latency_grb}), eight artifacts, one asteroid, and 38 with red stellar counterparts identified as M dwarfs.
+\textcolor{red}{Their Fig 1 is a delta-mag vs.
+delta-time plot comparing afterglows to Mdwarf flares that shows how they can be photometrically distinguished (difference regions of that parameter space)}.
+\textcolor{red}{Existence of short duration M dwarf flares established; what's the science related to them?}
+
 \clearpage

### 2020-02-27 draft-hostassoc.tex
added draft-latency

\documentclass[DM,lsstdraft,authoryear,toc]{lsstdoc}
% lsstdoc documentation: https://lsst-texmf.lsst.io/lsstdoc.html

% Package imports go here.
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
\usepackage{enumitem}

% Local commands go here.

% To add a short-form title:
% \title[Short title]{Title}
\title[Host Association]{Host Galaxy Association for {\tt DIAObjects}}

% Optional subtitle
% \setDocSubtitle{A subtitle}

\author{%
M.~L.~Graham et al.
}

\setDocRef{DMTN-TBD}
\date{\today}
% Optional: name of the document's curator
% \setDocCurator{The Curator of this Document}
\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-xxx}}

\setDocAbstract{%
This document argues that, in order to better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2020-02-26}{Inception.}{Melissa Graham}
}

\begin{document}

% Cite requirements using these macros.
% \lsrreq \ossreq \dmreq \reqparam

% Create the title page.
% Table of contents is added automatically with the "toc" class option.
\maketitle

% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Introduction} \label{sec:intro}

LSST will issue alert packets within $60$ seconds for all sources detected during difference image analysis (DIA; {\tt DIASources}), which are associated by sky coordinate into objects ({\tt DIAObject}). 
Individuals and brokers processing alerts will use the information in these packets to rapidly evaluate and prioritize {\tt DIAObjects} for follow-up with limited resources.
Thus the contents of the alert packet have been designed to contain a sufficient amount of LSST data about each {\tt DIAObject} to enable immediate analysis.

One important piece of information is the association of each {\tt DIAObject} with a static-sky {\tt Object} from the Data Release catalogs.
Brokers will use the {\tt Object} association to obtain data about the static-sky object from the DR catalogs, such as whether it might be galactic or extragalactic, at high- or low-redshift, nuclear or offset from a host, etc.
All of this information can help an alert stream user identify and prioritize their targets of interest, and delivering alerts \emph{with} the static-sky association already completed avoids the situation of multiple users cross-matching in real time.
The {\tt Object} association for {\tt DIAObjects} will of course also be used by scientists working with the Prompt or Data Release {\tt DIAObjects} catalogs on longer timescales, not just alert consumers.
However, the main goal of this document is to assess the best option -- from a scientific perspective -- for {\tt Object} association that can be completed during the 60 second Alert Production timescale.

Below, Section \ref{sec:DMplans} describes the current plan for associating {\tt DIAObjects} with the DR {\tt Objects} catalog; Section \ref{sec:options} presents and discusses options that have been used successfully by other surveys; and Section \ref{sec:recs} makes recommendations for additions to the {\tt DIAObject} catalog to improve associations for extragalactic transients and their host galaxies.

\section{Current DM Plans for Host Galaxy Association}\label{sec:DMplans}

With respect to associations between Prompt {\tt DIAObjects} and Data Release {\tt Objects}, the contents of the alert packet as defined in \citeds{LSE-163} includes the following:
\begin{itemize}
\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
\end{itemize}
For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.

The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxies' spatial luminosity profiles.
Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".

\section{Options to Improve Host Galaxy Association}\label{sec:options}

Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
This is usually estimated by calculating a \emph{separation distance} from the nearby galaxies to the transient which is expressed in terms of the galaxy's spatial luminosity profile, and then assuming the galaxy with the lowest separation distance is the host.
The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.

\subsection{Effective Radius}

For the separation distance, use the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy.
The DR {\tt Objects} table is already planned to contain suitable effective radii such as the parameter {\tt kronRad90} \citedsp{LSE-163}.
This option for the separation distance would not require any additional processing aside from dividing the radial distance from transient to {\tt Object} by the effective radius of the {\tt Object}.
Although this kind of separation distance would account for the relative sizes of the potential host galaxies, it does not account for their position angles, and so would not be as accurate for assessing potential high-inclination host galaxies.

\subsection{Second Moments}

Calculate a separation distance based on the two-dimensional luminosity profile of the nearby galaxies.
For example, \citet{2006ApJ...648..868S} describe the method applied to the Supernova Legacy Survey (SNLS), using a separation distance of $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $C_{xx}$, $C_{yy}$, and $C_{xy}$ are ellipse parameters derived from the second moments of the galaxy luminosity profile and $x_r$,$y_r$ are the on-sky distances between the centroids of the transient and the galaxy.
The DR {\tt Objects} table is already planned to contain the second moments of the galaxy luminosity profiles ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy}; \citeds{LSE-163}).
A step-by-step description of how this separation distance can be calculated from planned {\tt DIAObject} and {\tt Object} table elements is provided in Section \ref{sec:app}.

Multiple recent surveys have used this method (or similar) to associate transients with their host galaxies, such as \citet{2018PASP..130f4002S} for SDSS supernovae, and \citet{2016AJ....152..154G} who use real and simulated data to evaluate the optimal method for host association.
This option for the separation distances requires slightly more computational steps, but would account for both the relative sizes and position angles of the potential host galaxies.

\subsection{2D Algorithms}

Aside from adopting a separation distance, there are more complicated methods for identifying the most likely host for a given transient.
For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and that can estimate the fraction of light in given pixel that should be attributed to each (e.g., as the SCARLET deblender can do, \cite{2018A&C....24..129M}).
The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.
While all DR {\tt Objects} will be associated with a footprint (a region of connected pixels), the footprint information will not be stored in the {\tt Object} table.
The use of footprints in identifying potential host galaxies would require more computational resources during Prompt processing, but would probably return a more accurate host association for only a very small fraction of {\tt DIAObjects}.

\subsection{Hostless Transients}

For some scientific analyses, transients which are $>$3-5 effective radii away from the nearest galaxy, or for which $>99\%$ of the potential host's luminosity is within the separation distance, are considered ``hostless" (e.g., \citealt{2011ApJ...729..142S}).
Such a cutoff has been appropriate for past samples of $\sim$hundreds of transients, but will not be appropriate for the LSST sample size.
Furthermore, the decision of whether and how to consider a transient ``hostless" is best left as a scientific decision for the end-user.
Thus, no such cut should be applied during the association of {\tt DIAObjects} and {\tt Objects}, and the most likely hosts should still be reported, even if the probability is low.

\subsection{Galaxy/Transient Types}

The association of transients with their host galaxy can be more accurate if their properties are also considered.
For example, the potential host galaxy's redshifts can be used to calculate separation distances in physical units, or to estimate the absolute brightness of the transient and consider whether it is physically plausible.
Priors based on the established correlations between transient types and host galaxy morphology or color can also be used to refine a probabilistic host association, such as how core collapse supernovae are almost always associated with star formation (except for a few notable cases, e.g., \citealt{2012ApJ...753...68G,2019ApJ...887..127I}).
A demonstration that these correlations between host and transient types are so robust that the host type can be used to provide a statistical classification of the transient type was presented by \citet{2013ApJ...778..167F}.

% At DESC, Alex Gagliano presented their work on host association using PCA on teh galaxy properties, which shows that color and morphology are the first two PCA components, and that the hosts of SNIa/SNII etc. occupy different parts of the PC1 vs PC2 plane. That paper is still in prep. but could maybe be cited here eventually.

However, making science-informed associations between transients and galaxies based on any kind of derived properties is beyond the scope of Prompt processing, and is best left to the users on a case-by-case basis. 
Thus, properties of the transients and/or the nearby galaxies (beyond their coordinates and luminosity profile) should not be used during the association of {\tt DIAObjects} and {\tt Objects}.
For a very thorough assessment of an optimized, science-driven system for associating supernovae and their host galaxies -- including the role and performance of machine learning methods -- we direct the reader to \citet{2016AJ....152..154G}.

\section{Recommendations}\label{sec:recs}

For the ten\footnote{This is a conservative estimate.} {\tt Object} catalog galaxies (stars excluded) that are nearest to a given {\tt DIAObject} in terms of radial distance, a separation distance should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile (as described in Section \ref{sec:app}).

Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the objectId for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances for those three galaxies.

An analog for the existing element {\tt nearbyObjLnP}, which represents the probability of association for static but variable point sources (stars), is not necessary for potential host galaxies. The existing {\tt DIAObject} catalog elements {\tt nearbyObj} and {\tt nearbyObjDist} can remain unchanged.

This would add unit64[3] and float[3] to the {\tt DIAObject} catalog and to each alert ? a small and worthwhile addition.

\subsection{Draft RFC}\label{ssec:draft_rfc}

\textit{The following text should be posted as a Request For Comments (RFC) in Jira, and at the same time this DMTN should be made official and available.}

In order to better enable extragalactic transient science with brokers, it is proposed that two new {\tt DIAObject} table elements be computed during Alert Production:
(1) {\tt nearbyPotHost}, containing the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest \emph{separation distances}, and
(2) {\tt nearbyPotHostSepDist}, the separation distances for those three {\tt Objects}.
The \emph{separation distance} should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile, as described in detail in Section \ref{sec:app} of DMTN-XXX.
All parameters required for the calculation of the separation distances are already planned to be in the {\tt DIAObject} and {\tt Object} tables, and this change would only add unit64[3] and float[3] per {\tt DIAObject} catalog entry, and per alert.

\clearpage
\section{Appendix: Separation Distance from Second Moments}\label{sec:app}

From the LSST catalogs the following table elements are used to define the parameters needed to calculate the separation distance \citedsp{LSE-163}:
\begin{center}
\begin{tabular}{ccll}
\hline
Parameter & Unit & Table Element & Description \\
\hline
$x_{\rm trans},y_{\rm trans}$ & $\rm degrees$ & {\tt DIAObject} {\tt radec} & transient centroid  \\
$x_{\rm gal},y_{\rm gal}$       & $\rm degrees$ & {\tt Object} {\tt radec}       & galaxy centroid      \\
$\overline{x^2}$,$\overline{y^2}$,$\overline{xy}$ & $\rm arcsec^2$  & {\tt Object} {\tt Ixx}, {\tt Iyy}, {\tt Ixy} & galaxy second moments \\
\hline
\end{tabular}
\end{center}

As described in Section 10 of E. Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}} (and presumably many other places), the unitless ellipse parameters $C_{xx},C_{yy},C_{xy}$ can be calculated from the second moments via:
\begin{equation}
C_{xx} = \frac{\overline{y^2}}{\sqrt{ \left( \frac{\overline{x^2}-\overline{y^2}}{2} \right)^2 + \overline{xy}^2}}
\end{equation}
\begin{equation}
C_{yy} = \frac{\overline{x^2}}{\sqrt{ \left( \frac{\overline{x^2}-\overline{y^2}}{2} \right)^2 + \overline{xy}^2}}
\end{equation}
\begin{equation}
C_{xy} = -2 \frac{\overline{xy}}{\sqrt{ \left( \frac{\overline{x^2}-\overline{y^2}}{2} \right)^2 + \overline{xy}^2}}
\end{equation}

The sky distances between the transient and galaxy centroids are calculated as follows, and include the cos-dec factor and a conversion from units of degrees to arcseconds:
\begin{equation}
x_r = 3600(x_{\rm SN} - x_{\rm gal})
\end{equation}
\begin{equation}
y_r = 3600(y_{\rm SN} - y_{\rm gal})\cos{y_{\rm gal}}
\end{equation}

Finally, the separation distance $R$ in arcseconds is calculated as:
\begin{equation}
R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r.
\end{equation}

\clearpage
% Include all the relevant bib files.
% https://lsst-texmf.lsst.io/lsstdoc.html#bibliographies
\bibliography{local,lsst,lsst-dm,refs_ads,refs,books}

\end{document}

### 2020-02-27 draft-latency.tex
added draft-latency

--- 2020-02-27 draft-alertcontents.tex
@@
 % \title[Short title]{Title}
-\title[Alerts Menu]{LSST Alerts: Science-Driven Options for Packets and their Distribution}
+\title[Alerts Latency]{Science Drivers for LSST Alert Distribution Latency}
 
@@
 \author{%
-M.~L.~Graham and the Data Management System Science Team, with contributions from the Transients and Variable Stars Science Collaboration
+M.~L.~Graham, with contributions from the Transients and Variable Stars Science Collaboration
 }
@@
 \textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
-A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
+A discussion of the science drivers regarding the alert packet distribution latency.
 }
@@
 \setDocChangeRecord{%
-  \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
-  \addtohist{0.1}{2020-01-09}{First draft ready for input from others.}{Melissa Graham}
+  \addtohist{0}{2019-11-14}{Inception.}{Melissa Graham}
+  \addtohist{0.1}{2020-01-09}{Early draft.}{Melissa Graham}
 }
@@
 
-The purpose of this document is to consider the science motivation for, and impact of, various options for alert packet contents and their distribution timescale.
-The contents and delivery of alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public, as they are exempt from any proprietary period, and publicly accessible because they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
+This document considers the science motivation and impacts of the alert distribution timescale, also called the latency.
+Alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public (exempt from any proprietary period) and publicly accessible, as they are distributed to community brokers and not solely available via the LSST Science Platform (LSP).
 In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
 
-It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout.
-This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
-Reducing the packet size or increasing (slightly) the latency could enable the stream to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
-The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
-Reducing the contents should only be done if that information is easily and quickly obtainable elsewhere.
+It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout --- more specifically, that the data management system (DMS) be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
+The 5 full streams aspect is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility \citedsp{DMTN-102}.
+The 60 second aspect is scientifically motivated by a desire to explore the parameter space for very short timescale transients and variables which might last only a few minutes \citedsp{LPM-17}.
 
-{\bf Document Overview ---} \S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
-\S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
-\S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
+Increasing the alert latency could enable the stream to be delivered to more brokers, thereby increasing the amount of alerts-related science done by the community.
+Or, it might allow human and computational resources currently dedicated to building a DMS that can deliver a 1 minute latency to be repurposed. 
+In case it ever becomes technically necessary to change the alert latency, this document collects the science drivers for short timescale alert distribution.
+\textbf{Time-domain astronomy remains a rapidly evolving field and this should be considered a living, in-progress document}.
+
+\subsection{Summary}
+
+This document reports that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
 
@@
 %\begin{itemize}
-%\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
-%\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
-%\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
-%\item DM-20298, Revise the contents of the alert packets (Graham)
 %\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
-%\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
 %\end{itemize}
@@
 % % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Alert Packet Contents} \label{sec:packets}
+\section{Science Drivers for Alert Latency} \label{sec:latency}
 
-In this section the scientific impacts of various options to reduce the size of alert packets (in order to, e.g., enable wider alert distribution, facilitate broker processing) are considered, as are potential additions to the alert packet contents which would improve the user's ability to classify and prioritize alert sources for follow-up.
-The main conclusions of the following discussion are:
-\begin{itemize}
-\item {\bf Removing Histories (\S~\ref{sssec:packets_remove_hist}) ---} Removing the {\tt DIASource} histories from the alert packets that are transmitted to brokers that request this modification could be scientifically beneficial, but a list of all {\tt DIASource} identifiers associated with the {\tt DIAObject} at the time of alert generation must be added to those packets.
-\item {\bf Removing Stamps (\S~\ref{sssec:packets_remove_stamps}) ---} Removing the image stamps from all alert packets could be scientifically beneficial, but stamps should still be made during alert production and made publicly retrievable with a latency similar to that of alert distribution.
-\item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
-\item {\bf Adding Object Association Information (\S~\ref{ssec:packets_add}) ---} To better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
-\item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
-\end{itemize}
-
-\subsection{Removing Histories and/or Stamps}\label{ssec:packets_remove}
-
-The size of a fully-loaded individual alert packet is estimated to be $\lesssim$82~KB, based on simulations of the planned content of the alerts as described in Section 3.5 of \citeds{LSE-163} (see also \citeds{DMTN-102}).
-The two largest components of the alert packet which could be considered ``optional'' are the history and the postage stamps.
-The history is the past $\sim$12 months of {\tt DIASource} records, previously released as alerts.
-The postage stamps are, at minimum, 30$\times$30 pixels and contain flux, variance, and mask extensions for both the template and difference image, plus a header of metadata \citedsp{LSE-163}.
-The history accounts for $\sim$27KB of the alert packet ($\sim$33\%) and the stamps contribute $\gtrsim$18~KB ($\sim$20\%).
-
-Brokers that save all alerts, build their own a database of alerts contents, or have an automated interface with the LSST Prompt products database or Alerts database might not need the historical records to be included in the alert packets.
-Brokers which do not use the image stamps -- or would need them for only a subset of the alerts and could feasibly query a stamps database -- might not need them included in the alerts packets.
-Individual broker teams may indicate which information they require (or would like removed from the packets in their stream) during the broker proposal process \citedsp{LDM-612,LDM-682,LDM-723}.
-During the first stages of the proposal process, and at the LSST Community Brokers Workshop\footnote{\url{ls.st/cbw}} in June 2019, there was some indication that at least a few brokers were interested in alerts without historical records and/or postage historical records.
-
-\subsubsection{{\tt DIASource} Record History}\label{sssec:packets_remove_hist}
-
-Including the {\tt DIASource} record history in every alert allows for the \emph{instant assessment} --- without the need to cross-match or query catalogs --- of changes in the objects's location, size, shape, or brightness, provides context for the latest detection (i.e., anomalous or typical), and enables a robust assessment of the likely physical nature of the object (e.g., light curve fits indicating supernova type become more reliable with more observations).
-
-To remove the {\tt DIASource} record history from an alert would slightly inhibit the scientific assessment of transient/variable objects for brokers that build their own database of alert contents, due to the time required to cross-match with the database.
-Removing the {\tt DIASource} record history would completely inhibit this kind of analysis for brokers that do not have such storage or processing capabilities.
-
-There is also the issue that the association of {\tt DIASources} into {\tt DIAObjects} may change over time, for some objects.
-For example, rare cases such as multiply-imaged strongly-lensed supernovae, or two unassociated transients/variables that are superimposed along the line of sight, might be blended/separated in poor/good seeing.
-In such cases, the set of all associated {\tt DIASources} might be improved between one alert and the next.
-
-{\bf Summary --} The {\tt DIASource} record histories should not be removed from all alert packets, but could be optionally removed at the request of a broker.
-However, if the record histories are removed, a list of all {\tt DIASource} identifiers (the {\tt diaSourceId}, as in Section 3.3.1 of \citeds{LSE-163}) which are currently associated with the {\tt DIAObject} to which the alert-spawning {\tt DIASource} is associated should be provided (note that this list is \emph{not} currently part of the {\tt DIAObject} record).
-
-\subsubsection{Forced Photometry History}\label{sssec:packets_remove_fp}
-
-When forced photometry ({\tt DIAForcedSource} record history) is included in an alert packet, it is ``historical" in the sense that it is based on past images and not the image which spawned the alert.
-However, it is not always redundant information that was previously released in an alert, like the {\tt DIASource} record history.
-
-There are two main cases where forced photometry is performed, as described in \citeds{LSE-163}.
-(1) For all new {\tt DIAObjects}, forced ``pre-covery'' photometry is done in the past 30 days of difference images within 24 hours, stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
-(2) For all {\tt DIAObjects} that are not detected\footnote{The detection threshold is a signal-to-noise ratio $\geq$ 5.} in a given difference image, but had a detection in the past $\sim$12 months, forced photometry is done on that difference image.
-The result is stored in the {\tt DIAForcedSource} catalog, and included in all future alert packets associated with that {\tt DIAObject}.
-The forced photometry is scientifically valuable because, for example, pre-covery forced photometry provides context for the first detection of a transient/variable, and low-SNR detections can help with photometric classification and prioritization for follow-up observations.
-
-If a given {\tt DIAObject} is never again detected in a difference image, then the {\tt DIAForceSource} records are never included in an alert packet, but are available to individuals or brokers with access to the PPDB via the LSST Science Platform.
-If a given {\tt DIAObject} is repeatedly detected in new difference images, then the same {\tt DIAForcedSource} records will be redundantly included in all future alerts (after the first one containing the pre-covery or non-detection forced photometry).
-For the ``historical" {\tt DIAForcedSource} records, alert content redundancy could be avoided by only including previously undistributed {\tt DIAForcedSource} data in an alert packet.
-To do this, the Prompt pipeline would have to either compare with the contents of the most recent alert and remove redundant data from the new alert, or flag the {\tt DIAForcedSource} data as ``released" and only add ``unreleased" forced photometry to a alert packet.
-Both of these options would add processing times to the alert generation pipeline, which is undesirable.
-
-{\bf Summary --} If the decision is made to remove the historical {\tt DIASource} records from alert packets, the {\tt DIAForcedSource} records of forced photometry should always be added to all alerts.
-This will cause some redundancy in alert contents, but if not included there would be no public access to this scientifically valuable information.
-
-\subsubsection{Postage Stamps}\label{sssec:packets_remove_stamps}
-
-\textcolor{red}{MLG notes from DESC: not all alerts need them so it could be, e.g., for new DIASources only, non-stellar DIASources, etc.
-with the rest going to a server that does not have the 1 minute latency requirements.}
-
-Including the postage stamps in the alert packet allows for the instant assessment of the {\tt DIASource} in the difference and template images, which has a variety of scientific applications.
-For example, although real/bogus classification of the difference-image sources will be done prior to alert generation, some brokers may still wish to run additional algorithms on the images -- especially in the case of first detections.
-Brokers might also employ algorithms to classify transients and variables based on the images instead of the photometry (e.g., \citealt{2019PASP..131j8006C}).
-From the postage stamp images, scientifically useful information can be derived about the trails left by moving objects, or from the 2-dimensional luminosity distribution of spatially evolving objects such as comets, which might not be entirely captured by the {\tt DIASource} elements (see Table 1 in \citeds{LSE-163}).
-The images also provide context for the {\tt DIASource}, such as host-galaxy morphology or field crowdedness, that can assist with follow-up observations.
-
-If postage stamps were not created during alert production, the only option would be for users to wait until the images are made available (within 24 hours) in the LSP, which would significantly inhibit alerts-based science goals.
-However, postage stamps are unlikely to be needed for most {\tt DIASources}, the majority of which --- 95\%, according to the breakdown in \citeds{DMTN-102} --- will be stars or asteroids: point sources for which context is not a significant part of their immediate evaluation.
-An efficient compromise might be for the Prompt pipeline to create the postage stamps during alert generation, and then store them in a publicly accessible database (i.e., outside the LSP) where they can be obtained (via {\tt diaSourceId}) with a latency similar to the alert distribution timescale by anyone, from anywhere, at anytime.
-
-The expected maximum download rate from this stamp database is estimated by assuming that all 5 community brokers download $\sim$500 stamp sets per visit: $\sim$2500 stamp sets every $\sim$35 seconds.
-At 18~KB per stamp set, that's a data rate of 10.5 Mbps (for comparison, that's $\sim$5\% of one full alert stream).
-
-{\bf Summary --} The postage stamps could be removed from all alert packets, but they should still be created during Prompt processing and stored in a publicly accessible database where they can be obtained with a latency similar to the alert distribution timescale.
-
-\textcolor{red}{{\bf Could Postage Stamps Be Smaller? --}}
-\citeds{LSE-163} describes the size of the postage stamp to be at least $6\arcsec \times 6\arcsec$ ($30\times30$ pixels), or large enough to encompass the entire footprint of the {\tt DIASource} (e.g., if trailed, extended).
-That is large enough to contain the angular diameter (30 kpc) of a Milky Way-like galaxy at a redshift of $z\approx0.5$, which will be a fairly typical supernova host galaxy in the LSST data set.
-Since context is a science driver for producing stamps, it would not be desirable for them to be much smaller than $6\arcsec \times 6\arcsec$.
-
-\subsubsection{Pros and Cons of Multiple Alert Formats}\label{sssec:packets_remove_procon}
-
-The histories and/or stamps could be removed from all alert packets, or only at the request of a broker.
-In the latter case, LSST might end up distributing four types of alert packet formats to its five brokers: (i) with histories and stamps, (ii) with histories but without stamps, (iii) without histories but with stamps, and (iv) without histories or stamps.
-
-From a science perspective, the benefits to allowing brokers to customize their alert packet is that it potentially enables unique science to be done by brokers (e.g., very rapid custom stamp-analysis algorithms), and could enable LSST to support $>$5 brokers which might lead to more alerts-related science in general.
-
-One drawback might arise during times of heavy load, such as many visits with $>$40,000 alerts (\S~\ref{sec:graceful}), \emph{if}, and only if, full-sized packet streams are more likely to be delayed.
-Whether this becomes a possibility will depend on the technical aspects of how the multiple types of alerts are generated and transmitted, and also on the final decisions about latency (\S~\ref{sec:latency}) and delayed distribution (\S~\ref{sec:graceful}).
-This may well never come to pass, but if it did, it could be unfair to brokers requiring full-size packets.
-While some brokers are being designed for a wide variety of science goals, others are more specifically tailored.
-If it is the specifically-tailored broker that requires full-sized alerts, then experiencing delayed transmission more frequently than other brokers could put some science goals at a disadvantage. 
-Furthermore, the fact that not all alert packets are identical might cause some additional bookkeeping considerations for downstream brokers who ingest from multiple community brokers, but that should not be detrimental to their science goals.
-The technical considerations of how many different alert packet formats it is feasible to create and transmit within the required latency time is beyond the scope of this document, in which we focus on science impacts.
-
-{\bf Summary --} Offering multiple alert formats to brokers is scientifically beneficial provided that full-sized alerts are not predisposed to delay.
-
-\subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}
-
-\textcolor{red}{MLG notes from DESC: \\ (1) Check "Host Galaxy Identification for Supernova Surveys" \citet{2016AJ....152..154G}.
-\\ (2) Alex Gagliano is working on host association as a part of classification that does a PCA on galaxy properties, and found that color and morphology are the first two PCA components, and that the hosts of SNIa SNII etc.
-occupy different parts of the PC2 vs PC1 plane.
-His paper is in prep still.
-\\ (3) As MWV was trying to tell you, host association depends on the redshift of the object (i.e., you want to do it in absolute distance); obviously that won't be in the alert so we can't make it part of the "3 nearest", but you'll probably want to mention that specifically.}.
-
-The contents of the alert packet as defined in \citeds{LSE-163} includes the following with respect to associations with {\tt Object} catalog from the most recent data release:
-\begin{itemize}
-\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
-\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
-\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
-\end{itemize}
-For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.
-
-The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
-For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxy's luminosity profile.
-Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".
-
-Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
-This is usually estimated by calculating an offset distance from the nearby galaxies to the transient that is expressed in terms of the galaxy's spatial luminosity profile, and assuming the galaxy with the lowest offset distance is the host.
-The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.
-
-{\bf Effective Radius --} Calculate a separation distance that is the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy (i.e., {\tt kronRad90} in the {\tt Objects} table; \citeds{LSE-163}).
-The nearby galaxy with the lowest separation distance is the most likely host. This will account for the relative sizes of the potential host galaxies, but not their position angles.
-
-{\bf Second Moment --} Calculate a separation distance, as in \citet{2006ApJ...648..868S}, based on the two-dimensional luminosity profile of the galaxy.
-Where $x_{\rm trans},y_{\rm trans}$ is the location of the transient, and $x_{\rm gal},y_{\rm gal}$ is the center of the galaxy (the first moment; {\tt radec} in the {\tt Object} table), the separation distance is $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $x_r = x_{\rm SN} - x_{\rm gal}$ and $y_r = y_{\rm SN} - y_{\rm gal}$.
-The ellipse parameters $C_{xx}$, $C_{yy}$, and $C_{xy}$ can be calculated directly from the second moments of the galaxy luminosity profile ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy} in the {\tt Object} table), e.g., as described in Section 10 of E.
-Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}}.
-The nearby galaxy with the lowest separation distance is the most likely host.
-This option accounts for both the relative sizes and position angles of the potential host galaxies, and requires only a little more processing.
-
-{\bf 2D Algorithms --} There are also other, more complicated methods for identifying the most likely host for a given transient.
-For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
-Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
-\textcolor{red}{\citeds{LSE-163} does mention that footprints will be generated for {\tt Objects}, but it doesn't look like they'd be stored in the {\tt Object} table; find out if these footprints would be appropriate/available to use here.}
-The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.
-
-Generally, nearby galaxies with separation distances $>$3, or $>$99\% of the luminosity profile, are not assigned as hosts and the transient is considered "hostless".
-While this cutoff has been appropriate for past samples of $\sim$hundreds of transients, it will not be appropriate for the LSST sample size, and no such cut should be applied when evaluating potential host galaxies for {\tt DIAObjects}.
-
-All of the above host galaxy identification methods can also make use of priors if the transient type is known; for example, the established correlation between core-collapse supernovae and star formation.
-However, such a robust level of host associations are beyond scope for the Prompt pipeline, and the main goal is to provide brokers with sufficient information to assess the potential host association, and decide whether it is useful to the classification and/or prioritization of alerts for follow-up.
-
-{\bf Summary --} For the ten nearest {\tt Object} catalog galaxies, a separation distance should be calculated with respect to the transient location, preferably using the second moments of each galaxy's luminosity profile.
-Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances.
-An analog for the existing element {\tt nearbyObjLnP}, representing the probability, is not necessary.
-This would add {\tt unit64[3]} and {\tt float[3]} to the {\tt DIAObject} catalog and to each alert, but this both tiny and worthwhile.
-
-\subsection{Compressing the Alert Packets}\label{ssec:packets_compress}
-
-The application of gzip compression could further reduce the size of a full alert to $\sim$65~KB (80\%; JIRA ticket DM-16280).
-Naively, this seems like it would allow 1 more full stream to be transmitted to a broker, and thereby enable more science.
-However, the time and computational resources required to compress the alert packets needs to be considered.
-For example, gzip compression at 50 MB/s to compress $\sim$10000 alerts would take $\sim$10000 $\times$0.08 MB per alert $/$50 MB/s, or $\sim$16 seconds.
-This is significant, considering the alert distribution latency requirement\reqparam{OTT1}\lsrreq{0101}\ossreq{0127}\dmreq{0004} is 60 seconds. 
-Furthermore, compressing the alert packets forces brokers to then decompress the alerts on arrival, which would incur further delay.
-Science goals requiring very low-latency alerts distribution might be negatively impacted by compression -- see \S~\ref{sec:latency} for a deeper discussion of the science drivers for low-latency alerts.
-
-{\bf Summary --} Since packet compression could induce an additional latency, perhaps it should only be done if, e.g., an algorithm could provide a compression rate of $\lesssim$10 MB/s (requiring only a few seconds to compress a visit's worth of alerts), or if the alert distribution latency requirement is relaxed from 1 minute (see \S~\ref{sec:latency}).
-Otherwise, compression would have no impact on alerts-related science goals and so this should be primarily a technically-driven decision.
-
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Alert Distribution Latency} \label{sec:latency}
-
-This section considers the scientific motivation for the 1 minute alert distribution latency, and the impact to alerts-related science if that latency was extended to, e.g., 5 minutes.
-It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
-However, if it is possible to relax this requirement without a negative impact on science then it might be possible for LSST to transmit alerts to $>$5 brokers (and/or repurpose human and computational resources dedicated to the 1 minute latency), which in turn might enable more alerts-related science.
-
-Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
+Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, their need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
 Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
@@
 
-After a consideration of the following science cases that are most likely to require rapid follow-up (summary below), it appears that most of them would be satisfied without risk if the latency was increased from 1 to 5 minutes.
-However, it is absolutely not unambiguously clear that a need for 1 minute alerts will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events over the next couple of years.
 \begin{itemize}
@@
 
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Pre-Filtered Alert Streams} \label{sec:prefilter}
-
-This section explores the scientific motivation and impact of offering brokers pre-filtered alert streams. 
-Pre-filtering could reduce the overall bandwidth of alert distribution and potentially allow streams to be transmitted to more brokers, thereby enabling more alerts-related science.
-Providing pre-filtered streams could also lessen the load and resources used by the brokers, enabling more alerts-related science by allowing brokers to spend a smaller fraction of their budget computational resources.
-
-Depending on the science goals of a given broker and/or their users, a broker might find it useful to only receive alerts for {\tt DIASources} which:
-\begin{itemize}
-\item are likely to have a certain physical origin, e.g., stellar variable, moving objects
-\item could potentially be observed by the follow-up resources its users have access to
-\item have at least one prior detection by LSST (because, e.g., their algorithms all require photometric confirmation or a change in brightness)
-\item originate in images obtained as part of a special program such as the deep drilling fields or target-of-opportunity imaging surveys
-\item were not obtained during poor weather conditions (e.g., an image quality or sky background violating some limit)
-\end{itemize}
-
-For the above science drivers, the types of pre-filters that are likely to be scientifically useful -- some of which have been suggested by brokers already -- include:
-\begin{itemize}
-\item {\bf Apparent Magnitude:} only {\tt DIASources} brighter than a specified apparent magnitude (for each of the LSST filters {\it ugrizy}) are transmitted. 
-\item {\bf Region of Sky:} only {\tt DIASources} in a given sky area (or areas; defined by, e.g., limits in right ascension or declination), are transmitted. 
-\item \texttt{\bf DIAObject} {\bf History:} only {\tt DIASources} associated with {\tt DIAObjects} that have at least $N$ previous detections are transmitted, where $N$ can be $\geq1$. 
-\item \texttt{\bf SSObject} {\bf Association:} only {\tt DIASources} that are associated with a {\tt SSObject} are transmitted.
-(Potentially including new, unassociated {\tt DIASources}, because they might also be moving objects).
-\item \texttt{\bf Object} {\bf Association:} only {\tt DIASources} associated with {\tt DIAObjects} that are, in turn, associated with a data release {\tt Object} that meets some specification are transmitted (e.g., associated with a stellar point source).
-\item {\bf Observation Metadata:} only alerts from images obtained with a given set of metadata, e.g., a special observing program or meeting sky condition thresholds, are transmitted.
-\end{itemize}
-
-All of these potential science-driven pre-filters are based on the information contained in the alert itself, and do not impose any additional processing on the Prompt pipeline.
-They all also have the potential to considerably reduce the number of transmitted alerts.
-Due to the exponential relationship between the number of variable stars and their variability amplitude, and also that of volume and distance modulus, an apparent magnitude limit $\sim1$ mag brighter than the nominal $5{\sigma}$ {\tt DIASource} detection limit could reduce the alert stream data rate by $\gtrsim50$\%.
-Limiting to alerts that are associated with past detections or certain types of objects could similarly reduce the alert stream data rate.
-These types of pre-filers would reduce the number of alerts per visit for all visits -- the average bandwidth -- which might enable alert distribution to extend to more brokers and thus enable more alerts-related LSST science.
-Pre-filters based on their sky location or observational metadata would probably either deliver all or none of the alerts for a given visit, causing fluctuations in the bandwidth but not necessarily in a way that would allow for additional streams to be transmitted.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\subsection{LSST Alert Filtering Service}\label{ssec:prefilter_lafs}
-
-As described in \citeds{LSE-163} the LSST will provide an alert filtering service with limited capacity, by which individuals with LSST data rights and access may receive alerts via pre-defined filters and/or create and apply their own filters to the stream \citedsp{LPM-17,LSE-61}.
-It is a requirement that the LSST alert filtering service be able to support\reqparam{numBrokerUsers}\dmreq{0343} at least 100 simultaneous users, and capable of returning 20\reqparam{numBrokerAlerts}\dmreq{0343} full-sized alerts per visit per user.
-User-generated filters may be comprised of, e.g., a series of {\tt if} statements applied to the alert packet elements.
-There are no requirements on the latency of filter execution, which may depend on the complexity of the filter and available resources.
-
-From a science perspective, it would be beneficial to offer any and all of the pre-filtered streams listed above as options for user-generated filters.
-Individual users running small follow-up programs are more likely to have access to smaller-aperture telescopes and thus desire an initial apparent magnitude cut.
-It would save on compute resources to avoid a situation in which, e.g., 50 user-generated brokers all start with an {\tt if} statement to filter the stream to, e.g., $m_r < 20$ mag.
-
-% One type of simple filtering that individual users are more likely to want than brokers is ``watchlist" capability -- lists of individual objects (stars or galaxies) and offset radii that alerts are desired within.
-
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Graceful Degradation} \label{sec:graceful}
-
-The LSST specifications for the DMS require that it support the distribution\reqparam{transN}\lsrreq{0101}\reqparam{nAlertVisitAvg}\ossreq{0193}\reqparam{nAlertVisitPeak}\dmreq{0393} of at least 40,000 alerts per single standard visit, and that for visits producing $\leq$40,000 alerts \reqparam{sciVisitAlertDelay}\reqparam{sciVisitAlertFailure}\ossreq{0112}\dmreq{0392} no more than 1\% of them fail to have at least 98\% of its alerts distributed within 60 seconds of image readout.
-It is furthermore specified that alert distribution "degrade gracefully" beyond that limit, meaning that visits resulting in an excess of 40,000 of alerts should not cause any DMS downtime \citedsp{LSE-30,LSE-61}.
-
-This leaves the open question of what, from a science perspective, is the optimal way of dealing with delayed alerts that also meets the DMS specification of a ``graceful degradation"?
-We leave the aspects of technical implementation of a ``graceful degradation", such as distributing delayed alerts and alert archive storage access, for elsewhere and here just consider the science implications for a delayed timescale.
-
-{\bf Next-opportunity distribution via the alert stream.} As discussed in Section \ref{sec:latency}, there remains plenty of science goals that do not absolutely require alert distribution in 1 minute.
-Distributing delayed alerts via the stream would still enable plenty of science.
-The brokers might prefer to have delayed alerts clearly flagged to properly process them (e.g., some filtering and processing done by brokers might only be appropriate for alerts delivered within a given latency).
-
-{\bf Next-morning distribution via the alert stream.} This refers to an option to collect all delayed alerts during the night and then releasing them (perhaps on a new ``topic") in the morning after survey operations have ended for the night.
-From a science perspective this is not as useful as next-opportunity distribution, but if it is preferred for technical reasons it would enable more science than the option below.
-
-{\bf Do not distribute delayed alerts; send directly to archive.} There is no scientific merit in not distributing delayed alerts, and two further drawbacks: the alert archive update timescale is 24 hours\reqparam{L1PublicT}, significantly slower than next-morning distribution, and the alert archive might only be accessible by brokers with data rights since it is part of the LSST Data Facility.
-
-% MLG: commented out technical implementation aspects.
-%The retention period of alert packets in the alert distribution system is still to be determined, but given the alert stream data rate of ~800 GB/night it is currently reasonable to expect a retention period of ~7 days.
-% It is a requirement that all alerts be stored in an archival database and be available for retrieval\ossreq{0185}.
-%The term ``available for retrieval" applies to users with data rights and access to the LSST Science Platform.
-%Like all other Prompt data products, the alerts archive will be updated within 24\reqparam{L1PublicT}\lsrreq{0104} hours \citedsp{LSE-29}.
-% However, note that the requirements on user access to the alerts archive are not yet fully developed, and may be limited in terms of how alert packets may be queried (e.g., alert ID), or by the bulk download capabilities.
-
 % Include all the relevant bib files.

### 2020-02-28 draft-alertcontents.tex
intermediate edits

--- 2020-01-10 dmtn-tbd.tex
@@
 % \title[Short title]{Title}
-\title[Alerts Menu]{LSST Alerts: Science-Driven Options for Packets and their Distribution}
+\title[Alerts Contents]{LSST Alerts: Science-Driven Options for Packet Contents and Pre-Filtering}
 
@@
 \author{%
-M.~L.~Graham and the Data Management System Science Team, with contributions from the Transients and Variable Stars Science Collaboration
+M.~L.~Graham and the Data Management System Science Team
 }
@@
 \textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
-A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
+A review and discussion of the variety of options for alert packets contents, pre-stream filtering, and so forth.
 }
@@
 \setDocChangeRecord{%
-  \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
-  \addtohist{0.1}{2020-01-09}{First draft ready for input from others.}{Melissa Graham}
+  \addtohist{0}{2019-11-14}{Inception.}{Melissa Graham}
+  \addtohist{0.1}{2020-01-09}{First draft.}{Melissa Graham}
 }
@@
 
-The purpose of this document is to consider the science motivation for, and impact of, various options for alert packet contents and their distribution timescale.
-The contents and delivery of alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public, as they are exempt from any proprietary period, and publicly accessible because they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
+The alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public (exempt from any proprietary period) and publicly accessible, as they are distributed to community brokers and not solely available via the LSST Science Platform (LSP).
 In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
+
+This document considers options for the alert packet contents and potential ``pre-filters" of the alert stream based on those contents. 
+The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
 
@@
 This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
-Reducing the packet size or increasing (slightly) the latency could enable the stream to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
-The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
-Reducing the contents should only be done if that information is easily and quickly obtainable elsewhere.
 
-{\bf Document Overview ---} \S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
-\S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
+Reducing the packet size by narrowing down the contents included, and/or reducing the number of alerts distributed by applying a pre-filter, could enable alerts to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
+
+\S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
 \S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
@@
 \item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
-\item {\bf Adding Object Association Information (\S~\ref{ssec:packets_add}) ---} To better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
 \item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
@@
 
-\subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}
-
-The contents of the alert packet as defined in \citeds{LSE-163} includes the following with respect to associations with {\tt Object} catalog from the most recent data release:
-\begin{itemize}
-\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
-\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
-\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
-\end{itemize}
-For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.
-
-The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
-For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxy's luminosity profile.
-Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".
-
-Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
-This is usually estimated by calculating an offset distance from the nearby galaxies to the transient that is expressed in terms of the galaxy's spatial luminosity profile, and assuming the galaxy with the lowest offset distance is the host.
-The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.
-
-{\bf Effective Radius --} Calculate a separation distance that is the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy (i.e., {\tt kronRad90} in the {\tt Objects} table; \citeds{LSE-163}).
-The nearby galaxy with the lowest separation distance is the most likely host. This will account for the relative sizes of the potential host galaxies, but not their position angles.
-
-{\bf Second Moment --} Calculate a separation distance, as in \citet{2006ApJ...648..868S}, based on the two-dimensional luminosity profile of the galaxy.
-Where $x_{\rm trans},y_{\rm trans}$ is the location of the transient, and $x_{\rm gal},y_{\rm gal}$ is the center of the galaxy (the first moment; {\tt radec} in the {\tt Object} table), the separation distance is $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $x_r = x_{\rm SN} - x_{\rm gal}$ and $y_r = y_{\rm SN} - y_{\rm gal}$.
-The ellipse parameters $C_{xx}$, $C_{yy}$, and $C_{xy}$ can be calculated directly from the second moments of the galaxy luminosity profile ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy} in the {\tt Object} table), e.g., as described in Section 10 of E.
-Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}}.
-The nearby galaxy with the lowest separation distance is the most likely host.
-This option accounts for both the relative sizes and position angles of the potential host galaxies, and requires only a little more processing.
-
-{\bf 2D Algorithms --} There are also other, more complicated methods for identifying the most likely host for a given transient.
-For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
-Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
-\textcolor{red}{\citeds{LSE-163} does mention that footprints will be generated for {\tt Objects}, but it doesn't look like they'd be stored in the {\tt Object} table; find out if these footprints would be appropriate/available to use here.}
-The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.
-
-Generally, nearby galaxies with separation distances $>$3, or $>$99\% of the luminosity profile, are not assigned as hosts and the transient is considered "hostless".
-While this cutoff has been appropriate for past samples of $\sim$hundreds of transients, it will not be appropriate for the LSST sample size, and no such cut should be applied when evaluating potential host galaxies for {\tt DIAObjects}.
-
-All of the above host galaxy identification methods can also make use of priors if the transient type is known; for example, the established correlation between core-collapse supernovae and star formation.
-However, such a robust level of host associations are beyond scope for the Prompt pipeline, and the main goal is to provide brokers with sufficient information to assess the potential host association, and decide whether it is useful to the classification and/or prioritization of alerts for follow-up.
-
-{\bf Summary --} For the ten nearest {\tt Object} catalog galaxies, a separation distance should be calculated with respect to the transient location, preferably using the second moments of each galaxy's luminosity profile.
-Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances.
-An analog for the existing element {\tt nearbyObjLnP}, representing the probability, is not necessary.
-This would add {\tt unit64[3]} and {\tt float[3]} to the {\tt DIAObject} catalog and to each alert, but this both tiny and worthwhile.
-
 \subsection{Compressing the Alert Packets}\label{ssec:packets_compress}
@@
 Otherwise, compression would have no impact on alerts-related science goals and so this should be primarily a technically-driven decision.
-
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Alert Distribution Latency} \label{sec:latency}
-
-This section considers the scientific motivation for the 1 minute alert distribution latency, and the impact to alerts-related science if that latency was extended to, e.g., 5 minutes.
-It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
-However, if it is possible to relax this requirement without a negative impact on science then it might be possible for LSST to transmit alerts to $>$5 brokers (and/or repurpose human and computational resources dedicated to the 1 minute latency), which in turn might enable more alerts-related science.
-
-Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
-Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
-For the purposes of this discussion, an optimistic outlook is assumed, and the difficulty in identifying targets with a single epoch of LSST photometry is not considered as a reason to downgrade the importance of a 1 minute alert latency.
-
-After a consideration of the following science cases that are most likely to require rapid follow-up (summary below), it appears that most of them would be satisfied without risk if the latency was increased from 1 to 5 minutes.
-However, it is absolutely not unambiguously clear that a need for 1 minute alerts will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events over the next couple of years.
-\begin{itemize}
-\item {\bf A TVS Survey (\S~\ref{ssec:latency_tvs})} found that only $<$10\% of respondents reported that requiring alerts within 1 minute was necessary for their science goals.
-\item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger).
-However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
-\item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.
-\item {\bf Fast Radio Bursts (\S~\ref{ssec:latency_frb}) ---} \textcolor{red}{Tentatively,} some models predict that a short, bright burst of optical emission from the physical mechanism powering FRBs could be detected by LSST 5 to 15 minutes prior to the radio signal.
-If such optical events occur with significant frequency, are bright and consistently detectable, and can be identified with a single-filter detection, and if radio telescopes can respond within minutes, then LSST is the most promising future facility for this science, which requires a 1 minute alert distribution latency.
-\item {\bf Young Supernovae (\S~\ref{ssec:latency_ysne}) ---} The science goals require follow-up in an hour or longer, and a 5 minute latency would not have a negatively impact.
-\item {\bf Solar System Objects (\S~\ref{ssec:latency_sso}) ---} \textcolor{red}{TBD.}
-\item {\bf Stellar Variables (\S~\ref{ssec:latency_stars}) ---} \textcolor{red}{TBD.}
-\end{itemize}
-
-\subsection{TVS Survey Results Regarding Alerts Timescales}\label{ssec:latency_tvs}
-{\it Contents contain the results of a TVS survey administered by Rachel Street.}
-
-The Transients and Variable Stars (TVS) science collaboration surveyed the science needs of their members with respect to alert latency, and the results\footnote{Results courtesy of TVS co-chair Rachel Street.} are shown in Table \ref{tab:tvs}.
-The survey asked respondents to choose the maximum tolerable and ideal average delay between the alerts being produced by the LSST data reduction pipeline and the alert information becoming available through the broker service.
-This is not exactly the same as the alert distribution timescale {\tt OTT1}, but these responses will inform the need for $<$1 minute alert distribution.
-Respondents were asked to provide a short summary of their science goals for alerts if they reported needed access within 15 minutes.
-Note that the pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.
-
-\begin{table}[h]%[htdp]
-\caption{Table of results from a TVS survey which asked "how fast do you really need alerts?".
-The total number of respondents was 20.
-The science driver acronyms are: EM-GW (electromagnetic counterparts to gravitational wave events), YSNe (young supernovae, including e.g., shock breakouts), GRBs (gamma-ray bursts).
-\label{tab:tvs}}
-\begin{center}
-\begin{tabular}{|l|cl|cl|}
-\hline
-             & Maximum & Science & Ideal       & Science \\
-Latency & Tolerable  & Driver(s) &  Average & Driver(s) \\
-\hline
-1 min or less & 1 & EM-GW, GRB  & 2 & EM-GW, YSNe, GRB \\
-1-5 min         & 0 &                          & 1 &                                   \\
-5-10 min       & 1 & EM-GW, YSNe & 1 & YSNe \\
-10-30 min     & 2 & YSNe               & 2 & EM-GW, GRBs \\
-30-60 min     & 2 &                          & 7 & EM-GW  \\
-1-6 hours     & 4 & EM-GW, GRBs & 1 &  \\
-6-12 hours   & 2 &                          & 1 &  \\
-12-24 hours & 3 & EM-GW            & 1 & EM-GW \\
-1-3 days      & 1 &                           & 0 &  \\
->3 days       & 4 &                           & 4 &  \\
-\hline
-\end{tabular}
-\end{center}
-\label{default}
-\end{table}%
-
-Only 20\% (10\%) of respondents report that $<$10 minutes ($\leq$1 minute) is an ideal average delay, and whereas 70\% report that $>$30 minutes would be a sufficient average latency.
-The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
-The fact that these three are also listed as science drivers for longer-latency alert is mainly due to the diversity within the science drivers and the fact that the events have both short- and longer-timescale features.
-Each of these science drivers is discussed in turn below, along with two other science cases that rely on rapid access to LSST alerts: fast radio bursts (FRBs) and solar system objects (SSOs).
-
-\subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
-{\it Some contents revised from a Slack conversation between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}
-
-During LSST Operations, target-of-opportunity imaging follow-up sequences might be executed in the error ellipse of a gravitational wave (GW) detection to search for the electromagnetic (EM) optical counterpart.
-Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
-Although the kilonova light curves last only a few days, since there is so far only one event, GW170817, a day or two delay between the GW event and the optical detection still yields very scientifically valuable data -- for now.
-During LSST Operations there will already exist a sizable collection of longer-latency follow-up, and it is likely that science will be moving in the direction of pushing to ever earlier detections.
-However, since kilonovae produce days-long optical afterglows, a 5 minute LSST alerts would likely suffice.
-
-There are two theoretical predictions for prompt optical emission that would require very rapid access to alerts.
-One of them is a potential faint, $<$1 hour, UV/optical transient that occurs at the time of jet-break out for a short gamma-ray burst associated with a binary neutron star merger (\S~\ref{ssec:latency_grb}).
-Another that predicts emission of a similar color, luminosity, and timescale is the spin-down energy of a long-lived ($10^2$--$10^4$ s) neutron star formed from a binary neutron star merger before its eventual collapse to a black hole \citep{2016ApJ...819...15S}.
-As described in \S~\ref{ssec:latency_grb}, very few events could be detected by serendipitous coincidence by the LSST WFD main survey, and targeted follow-up of well-localized GW events with more appropriate facilities such as space-based UV/optical imagers is much more likely to yield detections of this very short lived emission.
-
-There are two additional issues related to ToO for EMGW events which might make 1 minute alerts necessary or impossible.
-
-First, for rapid ToO observations when visit coordinates are not known in advance, there might not be time to pre-cache the slow database queries that are needed during alert production processing (e.g., loading slices of the PPDB catalogs).
-This might mean that an alert distribution latency cannot be guaranteed for ToO imaging surveys. 
-For the purposes of this science-driven discussion, assume this to be a surmountable technical implementation issue.
-
-The second is that, at least for run O3, the GW event detection system itself issues preliminary alerts within 1--10 minutes, and these preliminary alerts are often retracted and do not always have the sky localization.
-Even now, many imaging follow-up surveys wait for the initial alert (or retraction) to be sent after a round of human vetting of the GW event signal, and this can take several hours.
-With such latencies, the question of 1 {\it vs.} 5 minute LSST alert timescales becomes inconsequential.
-
-\subsection{Gamma-Ray Bursts}\label{ssec:latency_grb}
-
-{\it \textcolor{red}{MLG: no outside input on this yet, is needed to be confirmed by a GRB person.} }
-
-Short gamma-ray bursts (sGRBs) are thought to be the mergers of two neutron stars, or a neutron star and a black hole.
-As the jet propagates through the ejecta material a hot cocoon is formed.
-When cocoon and jet break out, along with the sGRB can be observed a blue optical transient with an absolute peak brightness of -12 to -15 magnitudes which lasts for $10^3$--$10^4$ seconds (e.g., \citealt{2018MNRAS.473..576G}).
-For an LSST detection limit of $r\sim24$ mag, a detection limit of $r\lesssim-12$ ($r\lesssim-15$) mag corresponds to distances of $\lesssim160$ ($\lesssim630$) Mpc, or redshifts $z\lesssim0.035$ ($z\lesssim0.14$).
-Observing the diversity of cocoon emission for a sample of sGRBs would require regular and consistent multi-band follow-up within $10^3$ seconds, or $15$ minutes.
-In such cases where a sGRB is observed in a time/area that is serendipitously coincident LSST visit, science would be maximized if the alert confirming the presence of an optical counterpart was distributed to brokers within 1 minute.
-However, what is the expected frequency of this situation occurring during LSST Operations?
-
-% z=0.035 is V=0.014 Gpc3
-% z=0.14 is V=0.836 Gpc3
-% (0.836/0.014) * 4/yr = 240/yr
-The rate of short GRBs in the local volume ($<$200 Mpc) is estimated to be quite low, $<$4 $\rm yr^{-1}$ \citep{2019arXiv190800100M}.
-Scaled up to the co-moving volume within 630 Mpc, this rate is $\sim$240 $\rm yr^{-1}$, all-sky.
-In the baseline main survey about a sixth of the sky can be observed in a given night, which implies $\sim$1 sGRBs with detectable optical counterparts in the WFD survey per night for every $\sim$10 nights of observations.
-The probability that one of the 1000 visits per night will occur during the 5 minutes in which an alert would be useful for triggering rapid follow-up is $<10^{-2}$.
-Very roughly, we might expect this serendipitous coincidence to occur twice a year at most, or $\sim$20 times over the 10-year LSST main survey.
-
-% 10 visits per 5 minute block, or 100 "blocks" per night
-% 100x100 potential sky area/time options for the 100 blocks, x10 days
-% of the options, 100 blocks per night x 10 nights would be done
-% and then randomly draw just the 1 sGRB
-% scipy.stats.hypergeom(100000,1000,1)
-% 0 0.9900 
-% 1 0.0100 
-% 2 0.0000
-
-For discovering the optical counterparts of sGRBs, the LSST wide-fast-deep survey is simply not the most efficient future facility for the job.
-As described by \citet{2018MNRAS.473..576G}, a rapid search at the location of sGRBs with a UV satellite such as ULTRASAT would be ideal.
-
-%Triggering criteria might be:
-%new {\it u}- or {\it g}-band source with recent limits associated with a $<$200 Mpc galaxy (how many expected per night?)
-%and then very high priority if it's serendipidously associated with a GRB detection in the last 5 minutes
-
-\subsection{Fast Radio Bursts}\label{ssec:latency_frb}
-
-{\it Some contents based on text from and discussions with Tyler Pritchard.}
-
-One of the most likely -- but also most mysterious -- transients that might significantly benefit from the $60$ second alert timescale are fast radio bursts (FRBs): a millisecond long pulse of coherent emission in the GHz range.
-The emission is dispersed by the inter-galactic medium (IGM), such that the pulse's observed arrival time is frequency dependent.
-Observed FRB dispersion measures of $\rm{DM}\approx100$ to $1000$ $\rm pc\ cm^{-3}$ indicate that they originate at cosmological distances, with redshifts $z\approx0.1$ to $1$ \citep{2018Natur.562..386S}.
-
-If an optical counterpart is generated by this coherent emission, the time delay between the optical detection and the radio detection is on the timescale of minutes for frequencies $\nu < 500\ {\rm MHz}$, as shown in Figure \ref{fig:sci_frb}.
-This leaves open the possibility for triggering radio follow-up of an optical counterpart candidate serendipitously detected by LSST -- if such counterparts exist and are detectable.
-A millisecond-long event in the optical would have to be quite bright to be detected by LSST, but studies show that LSST detections are feasible \citep{2016ApJ...824L..18L}.
-\cite{2019ApJ...878...89Y} demonstrate that two theoretical sources for coherent optical emission from FRBs would not be detectable by LSST, but that inverse Compton scattering processes could lead to optical detections with LSST (e.g., from pulsars or masers).
-Searches for FRB counterparts in optical images that were serendipitously obtained (near-) coincidentally with an FRB detection have only just recently become possible thanks to current wide-field imaging surveys.
-No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.
-
-Since the rate of FRBs is estimated to be quite high (thousands per day; \citealt{2016MNRAS.460L..30C}) even after radio detection efficiency is factored in there could be many within the observed LSST volume.
-For example, if there are 1000 FRB occurring across the full sky every 24 hours, and LSST observes one sixth of the sky during a night's 10 hours of darkness, then there would be $\sim70$ FRB occurring in that area during that time.
-Assuming that both the FRB and the LSST visits are distributed randomly in time and location (although the visits are correlated, obviously), there's a $\sim6\%$ chance that in any given night one visit will be serendipitously coincident with the expected arrival of a short lived optical precursor to the FRB.
-If such a source could be recognized as an FRB optical counterpart within a minute, and radio follow-up could begin as soon as possible to detect the radio component and measure the time delay (Figure \ref{fig:sci_frb}).
-However, \textcolor{red}{it is (or rather, MLG is) currently unclear whether (a) there is strong theoretical support for such optical counterparts, and (b) radio telescopes can perform ToO follow-up on $<$15 minute timescales}.
-
-If FRBs are associated with superluminous supernovae (SLSNe) and young magnetars, then potentially an LSST alert regarding a change in behavior of known SLSNe could be used to trigger radio follow-up.
-However, this case is unlikely to be as sensitive to the LSST alert distribution timescale as the optical emission would be released over a longer time window \citep{2019arXiv191002036L}.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=15cm]{figures/frb_optical_delays.png}
-\caption{The delay time (for coherent emission) between the arrival of the optical and radio photons due to cosmological dispersion, as a function of the frequency of the radio detector.
-The {\it lowest} frequency bands of the Square Kilometer Array (SKA) and the Low-Frequency Array (LOFAR) are marked with vertical dotted and dash-dotted lines, and the 1 minute timescale for LSST alert distribution with a horizontal dashed line.
-The delay time is $\Delta t = k_{\rm DM}\,DM\,(\nu_{\rm low}^{-2} - \nu_{\rm high}^{-2})$, where $k_{DM}=4.149$ $\rm GHz^2\ pc^{-1}\ cm^3\ ms$, ${\rm DM}$ is the dispersion measure in $\rm pc\ cm^{-3}$, $\nu$ is the low and high frequency bands of the observation, and $\Delta t$ is in $\rm ms$.
-We have used the range of FRB dispersion measures as observed by \cite{2018Natur.562..386S}.
-\label{fig:sci_frb}}
-\end{center}
-\end{figure}
-
-\subsection{Young Supernovae}\label{ssec:latency_ysne}
-
-Photometric data obtained during the first few hours to days of a supernova can constrain the progenitor radius and circumstellar material in the immediate environment, which in turn provides information about the progenitor star, its system, and its final stages of evolution before the explosion.
-For example, \citet{2018Natur.554..497B} model the optical photometric data of the post-shock breakout cooling peak for a massive star's core collapse supernova to constrain the radius and mass of its outer envelope; \citet{2012ApJ...744L..17B} used the first day of photometric observations of a nearby Type Ia (thermonuclear detonation) supernova to show that their progenitor stars must be a compact degenerates; and rare ``blue bumps" in the first few days of a Type Ia supernova's light curve might betray the presence of a non-degenerate companion star \citep[e.g.,][]{2010ApJ...708.1025K,2017ApJ...845L..11H}.
-
-For young supernovae, follow-up observations within a couple of hours are required, and the related science goals are unlikely to be impacted if the alert latency was 5--10 minutes instead of 1 minute.
-
-\subsection{Solar System Objects}\label{ssec:latency_sso}
-
-\textcolor{red}{MLG: in Lynne's talk on the SSSC needs for alerts brokers at the CBW in June 2019, slide 4 says that a needed broker capability was "API or other 'push' notification (e.g., trailed detections) [minutes may count]".
-Follow-up on the SS use case for alerts on very short timescales.}
-
-\subsection{Stellar Variables}\label{ssec:latency_stars}
-
-\textcolor{red}{Return to the road map or science book? Surely there must be some stellar use-cases for 1 minute alerts. E.g., microlensing peaks can be real sharp.}
 

### 2020-02-28 draft-latency.tex
intermediate edits

@@
 \setDocAbstract{%
-\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
-A discussion of the science drivers regarding the alert packet distribution latency.
+\textcolor{red}{\bf EXTREME DRAFT. DO NOT CITE OR READ.}\\
+A gathering place for the science drivers related to alert packet distribution latency.
 }
@@
 
-This document considers the science motivation and impacts of the alert distribution timescale, also called the latency.
+This document gathers the scientific motivations for --- and impacts of --- the alert distribution timescale, also called the latency.
 Alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public (exempt from any proprietary period) and publicly accessible, as they are distributed to community brokers and not solely available via the LSST Science Platform (LSP).
-In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
+Although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.
 
 It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout --- more specifically, that the data management system (DMS) be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
-The 5 full streams aspect is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility \citedsp{DMTN-102}.
-The 60 second aspect is scientifically motivated by a desire to explore the parameter space for very short timescale transients and variables which might last only a few minutes \citedsp{LPM-17}.
+The 60 second latency was broadly scientifically motivated by a desire to explore the parameter space for very short timescale transients and variables which might last only a few minutes \citedsp{LPM-17}.
+The number of 5 full streams is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility \citedsp{DMTN-102}.
 
-Increasing the alert latency could enable the stream to be delivered to more brokers, thereby increasing the amount of alerts-related science done by the community.
-Or, it might allow human and computational resources currently dedicated to building a DMS that can deliver a 1 minute latency to be repurposed. 
-In case it ever becomes technically necessary to change the alert latency, this document collects the science drivers for short timescale alert distribution.
+Increasing the alert latency \emph{might} enable the stream to be delivered to more brokers, thereby increasing the amount of alerts-related science done by the community.
+Or it might allow human and computational resources that are currently dedicated to building a DMS which can deliver a 1 minute latency to be repurposed. 
+\textbf{There is currently no tabled proposal to increase the alert latency from 60 seconds}, but in case there ever is, this document collects the science drivers for short timescale alert distribution.
 \textbf{Time-domain astronomy remains a rapidly evolving field and this should be considered a living, in-progress document}.
@@
 \item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger).
-However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
+However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO imaging remains $\sim$tens of minutes.
 \item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.

### 2020-02-28 draft-alertcontents.tex
updated alertcontents

@@
 % \title[Short title]{Title}
-\title[Alerts Contents]{LSST Alerts: Science-Driven Options for Packet Contents and Pre-Filtering}
+\title[Alerts Contents]{LSST Alerts: A Scientific Perspective on Packet Contents and Pre-Filtering}
 
@@
 \setDocAbstract{%
-\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
-A review and discussion of the variety of options for alert packets contents, pre-stream filtering, and so forth.
+\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE OR READ.}\\
+\textcolor{red}{The DM-SST will return to this document once the broker full proposals have been received, as they will contain more information on these topics.}\\
+A review and discussion of the science motivation for, and impacts of, the variety of options for alert packets contents (e.g., histories, stamps) and pre-filtered streams (based on, e.g., apparent magnitude, {\tt Object} assocation).
+A brief discussion about the science impacts for the various ways of implementing a ``graceful degradation" of $>$40000 alerts per visit.
 }
@@
 This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
-
 Reducing the packet size by narrowing down the contents included, and/or reducing the number of alerts distributed by applying a pre-filter, could enable alerts to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
@@
 \S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).
-
-%This draft contains content pertaining to the following open tickets in the DM-SST epic "Studies Around Alerts" (assignees):
-%\begin{itemize}
-%\item DM-19484, Estimate distribution of alert packet size based on scientific considerations (Bellm)
-%\item DM-15654, Define policy for handling fields with $>$10K alerts (Bellm)
-%\item DM-20296, Investigate the possibility of providing filtered streams (Graham)
-%\item DM-20298, Revise the contents of the alert packets (Graham)
-%\item DM-20299, Scientific use-cases requiring $<$5 minute access (Graham)
-%\item DM-20300, Investigate options for expanding access to the alert stream (Bellm)
-%\end{itemize}
 
@@
 One drawback might arise during times of heavy load, such as many visits with $>$40,000 alerts (\S~\ref{sec:graceful}), \emph{if}, and only if, full-sized packet streams are more likely to be delayed.
-Whether this becomes a possibility will depend on the technical aspects of how the multiple types of alerts are generated and transmitted, and also on the final decisions about latency (\S~\ref{sec:latency}) and delayed distribution (\S~\ref{sec:graceful}).
+Whether this becomes a possibility will depend on the technical aspects of how the multiple types of alerts are generated and transmitted, and also on the final decisions about delayed distribution (\S~\ref{sec:graceful}).
 This may well never come to pass, but if it did, it could be unfair to brokers requiring full-size packets.
@@
 Furthermore, compressing the alert packets forces brokers to then decompress the alerts on arrival, which would incur further delay.
-Science goals requiring very low-latency alerts distribution might be negatively impacted by compression -- see \S~\ref{sec:latency} for a deeper discussion of the science drivers for low-latency alerts.
+Science goals requiring very low-latency alerts distribution might be negatively impacted by compression.
 
-{\bf Summary --} Since packet compression could induce an additional latency, perhaps it should only be done if, e.g., an algorithm could provide a compression rate of $\lesssim$10 MB/s (requiring only a few seconds to compress a visit's worth of alerts), or if the alert distribution latency requirement is relaxed from 1 minute (see \S~\ref{sec:latency}).
+{\bf Summary --} Since packet compression could induce an additional latency, perhaps it should only be done if, e.g., an algorithm could provide a compression rate of $\lesssim$10 MB/s (requiring only a few seconds to compress a visit's worth of alerts), or if the alert distribution latency requirement is relaxed from 1 minute (not currently planned).
 Otherwise, compression would have no impact on alerts-related science goals and so this should be primarily a technically-driven decision.
@@
 
-{\bf Next-opportunity distribution via the alert stream.} As discussed in Section \ref{sec:latency}, there remains plenty of science goals that do not absolutely require alert distribution in 1 minute.
-Distributing delayed alerts via the stream would still enable plenty of science.
+{\bf Next-opportunity distribution via the alert stream.} There are plenty of science goals that do not absolutely require alert distribution in 1 minute, and so distributing delayed alerts via the stream would still enable plenty of science.
 The brokers might prefer to have delayed alerts clearly flagged to properly process them (e.g., some filtering and processing done by brokers might only be appropriate for alerts delivered within a given latency).
@@
 
+\clearpage
 % Include all the relevant bib files.

### 2020-02-28 draft-latency.tex
updates

@@
 A gathering place for the science drivers related to alert packet distribution latency.
-}
+\textcolor{red}{\textbf{There is currently no tabled proposal to change the alert latency from 60 seconds}}, but in case there ever is, this document collects the science drivers for short timescale alert distribution.}
 
@@
 Or it might allow human and computational resources that are currently dedicated to building a DMS which can deliver a 1 minute latency to be repurposed. 
-\textbf{There is currently no tabled proposal to increase the alert latency from 60 seconds}, but in case there ever is, this document collects the science drivers for short timescale alert distribution.
+\textcolor{red}{\textbf{There is currently no tabled proposal to increase the alert latency from 60 seconds}}, but in case there ever is, this document collects the science drivers for short timescale alert distribution.
 \textbf{Time-domain astronomy remains a rapidly evolving field and this should be considered a living, in-progress document}.

### 2020-02-28 draft-alertcontents.tex
update

@@
 
-\S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
+\S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency.
 \S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).

### 2020-03-26 draft-hostassoc.tex
added DPDD reference to Sec 4.1
fixed typo
minor update to 4.1

@@
 
-This would add unit64[3] and float[3] to the {\tt DIAObject} catalog and to each alert ? a small and worthwhile addition.
+This would add unit64[3] and float[3] to the {\tt DIAObject} catalog and to each alert, a small and worthwhile addition.
 
@@
 
-\textit{The following text should be posted as a Request For Comments (RFC) in Jira, and at the same time this DMTN should be made official and available.}
+\textit{The following text should be posted as a Request For Comments (RFC) in Jira, and at the same time this DMTN should be made official and available.
+The RFC would spawn an LSST Change Request on the Data Products Definitions Document (exact proposed changes to DPDD to be written at that time.}
 

### 2020-04-21 draft-hostassoc.tex
stars excluded --> extended objects

@@
 
-For the ten\footnote{This is a conservative estimate.} {\tt Object} catalog galaxies (stars excluded) that are nearest to a given {\tt DIAObject} in terms of radial distance, a separation distance should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile (as described in Section \ref{sec:app}).
+For the ten\footnote{This is a conservative estimate.} {\tt Object} catalog galaxies (extended objects) that are nearest to a given {\tt DIAObject} in terms of radial distance, a separation distance should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile (as described in Section \ref{sec:app}).
 

### 2020-05-11 draft-hostassoc.tex
v2 draft

--- 2020-03-26 draft-hostassoc.tex
@@
 
-With respect to associations between Prompt {\tt DIAObjects} and Data Release {\tt Objects}, the contents of the alert packet as defined in \citeds{LSE-163} includes the following:
+With respect to associations between Prompt {\tt DIAObjects} and Data Release {\tt Objects}, the contents of the alert packet as defined in the Data Products Definitions Document (DPDD; \citeds{LSE-163}) includes the following:
 \begin{itemize}
@@
 The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
-For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxies' spatial luminosity profiles.
+For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the radial distance in arcseconds matters less than a \emph{separation distance} which accounts for the galaxies' spatial luminosity profiles.
 Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".
@@
 Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
-This is usually estimated by calculating a \emph{separation distance} from the nearby galaxies to the transient which is expressed in terms of the galaxy's spatial luminosity profile, and then assuming the galaxy with the lowest separation distance is the host.
+This is usually estimated by first calculating a \emph{separation distance} from the nearby galaxies to the transient which is expressed in terms of the galaxy's spatial luminosity profile, and then assuming the galaxy with the lowest separation distance is the host.
 The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.
 
-\subsection{Effective Radius}
+\subsection{Effective Radius}\label{ssec:options_Re}
 
@@
 The DR {\tt Objects} table is already planned to contain the second moments of the galaxy luminosity profiles ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy}; \citeds{LSE-163}).
-A step-by-step description of how this separation distance can be calculated from planned {\tt DIAObject} and {\tt Object} table elements is provided in Section \ref{sec:app}.
+A step-by-step description of how this separation distance can be calculated from planned {\tt DIAObject} and {\tt Object} table elements is provided in Section \ref{sec:appA}.
 
 Multiple recent surveys have used this method (or similar) to associate transients with their host galaxies, such as \citet{2018PASP..130f4002S} for SDSS supernovae, and \citet{2016AJ....152..154G} who use real and simulated data to evaluate the optimal method for host association.
-This option for the separation distances requires slightly more computational steps, but would account for both the relative sizes and position angles of the potential host galaxies.
+This option for the separation distance requires slightly more computational steps, but would account for both the relative sizes and position angles of the potential host galaxies.
 
@@
 
-For some scientific analyses, transients which are $>$3-5 effective radii away from the nearest galaxy, or for which $>99\%$ of the potential host's luminosity is within the separation distance, are considered ``hostless" (e.g., \citealt{2011ApJ...729..142S}).
+For some scientific analyses, transients which are $>$3-5 effective radii away from the nearest galaxy, or for which $>99\%$ of the potential host's luminosity is within the radial distance between transient and host center, are considered ``hostless" (e.g., \citealt{2011ApJ...729..142S}).
 Such a cutoff has been appropriate for past samples of $\sim$hundreds of transients, but will not be appropriate for the LSST sample size.
@@
 The association of transients with their host galaxy can be more accurate if their properties are also considered.
-For example, the potential host galaxy's redshifts can be used to calculate separation distances in physical units, or to estimate the absolute brightness of the transient and consider whether it is physically plausible.
+For example, the potential host galaxy's redshift can be used to calculate separation distances in physical units, or to estimate the absolute brightness of the transient and consider whether it is physically plausible.
 Priors based on the established correlations between transient types and host galaxy morphology or color can also be used to refine a probabilistic host association, such as how core collapse supernovae are almost always associated with star formation (except for a few notable cases, e.g., \citealt{2012ApJ...753...68G,2019ApJ...887..127I}).
@@
 
-For the ten\footnote{This is a conservative estimate.} {\tt Object} catalog galaxies (stars excluded) that are nearest to a given {\tt DIAObject} in terms of radial distance, a separation distance should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile (as described in Section \ref{sec:app}).
+\textbf{1: } Two new {\tt DIAObject} catalog elements should be added: {\tt potentialHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt potentialHostSeparation}, containing the separation distances for those three galaxies.
 
-Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the objectId for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances for those three galaxies.
+\textbf{2: } These two new {\tt DIAObject} catalog elements should use a separation distance calculated with respect to the transient location using the second moments of each galaxy's luminosity profile (as described in Section \ref{sec:appA}).
 
-An analog for the existing element {\tt nearbyObjLnP}, which represents the probability of association for static but variable point sources (stars), is not necessary for potential host galaxies.
-The existing {\tt DIAObject} catalog elements {\tt nearbyObj} and {\tt nearbyObjDist} can remain unchanged.
+These new catalog elements add unit64[3] and float[3] to the {\tt DIAObject} catalog and to each alert, which is a small and worthwhile addition.
+Including the three galaxies (instead of the two or one) with the lowest separation distances is motivated by the high-reward science case of correctly associating transients in the outskirts of large nearby host galaxies, as described in Appendix \ref{sec:appB}.
+This science case also requires that separation distances be initially calculated for all galaxies within $\sim$200 arcsec (i.e., $4R_e$ of a large nearby galaxy); how to identify this sub-set of galaxies in the vicinity (e.g.,.
+radial on-sky distance, HEALpix) is left as a technical implementation detail.
+Large bright nearby galaxies which are not in the {\tt Object} catalogs should have their coordinates and second moments imported so that they are included in this potential host association process.
 
-This would add unit64[3] and float[3] to the {\tt DIAObject} catalog and to each alert, a small and worthwhile addition.
+The existing {\tt DIAObject} catalog elements {\tt nearbyObj} and {\tt nearbyObjDist} can remain unchanged and an analog for the existing {\tt DIAObject} catalog element {\tt nearbyObjLnP}, which represents the probability of association for static but variable point sources (stars), is not necessary for potential host galaxies.
 
@@
 \textit{The following text should be posted as a Request For Comments (RFC) in Jira, and at the same time this DMTN should be made official and available.
-The RFC would spawn an LSST Change Request on the Data Products Definitions Document (exact proposed changes to DPDD to be written at that time.}
+The RFC would spawn an LSST Change Request on the Data Products Definitions Document.}
 
 In order to better enable extragalactic transient science with brokers, it is proposed that two new {\tt DIAObject} table elements be computed during Alert Production:
-(1) {\tt nearbyPotHost}, containing the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest \emph{separation distances}, and
-(2) {\tt nearbyPotHostSepDist}, the separation distances for those three {\tt Objects}.
-The \emph{separation distance} should be calculated with respect to the transient location using the second moments of each galaxy's luminosity profile, as described in detail in Section \ref{sec:app} of DMTN-XXX.
+(1) {\tt potentialHost}, containing the {\tt objectId} for the three {\tt Object} catalog extended sources with the lowest separation distances, and
+(2) {\tt potentialHostSeparation}, the separation distances for those three {\tt Objects}.
+The separation distance should be calculated with respect to the transient location using the second moments of each {\tt Object}'s luminosity profile, as described in DMTN-XXX.
 All parameters required for the calculation of the separation distances are already planned to be in the {\tt DIAObject} and {\tt Object} tables, and this change would only add unit64[3] and float[3] per {\tt DIAObject} catalog entry, and per alert.
 
+The change request on the DPDD would be to add two rows to {\tt DIAObject} Table (Table 3), and a footnote.
+
+\begin{tabular}{p{4.5cm} p{2cm} p{8cm}}
+{\tt potentialHost} & unit64[3] & Three extended {\tt Objects} with lowest separations in Data Release database$^{\rm FN}$. \\
+{\tt potentialHostSeparation} & float[3] & Separations of {\tt potentialHost}.   \\
+\end{tabular}
+
+$^{\rm FN}$ Separations should be calculated with respect to the transient location using the second moments of each {\tt Object}'s luminosity profile, as described in DMTN-XXX.
+
 \clearpage
-\section{Appendix: Separation Distance from Second Moments}\label{sec:app}
+\section{Appendix A: Separation Distance from Second Moments}\label{sec:appA}
 
-From the LSST catalogs the following table elements are used to define the parameters needed to calculate the separation distance \citedsp{LSE-163}:
+From the LSST catalogs the following {\tt Object} table elements are used to define the parameters needed to calculate the separation distance \citedsp{LSE-163}:
+
 \begin{center}
@@
 
+There might be an issue with using the {\tt Object} catalog second moments: the {\tt Ixx}, {\tt Iyy}, and {\tt Ixy} are defined with with respect to the local tract/patch and not sky coordinates (Jira DM-19519).
+Although the local tangent projection will probably work fine for this application of the second moments, this should be verified at the time of implementation.
+
 As described in Section 10 of E.
 Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}} (and presumably many other places), the unitless ellipse parameters $C_{xx},C_{yy},C_{xy}$ can be calculated from the second moments via:
+
 \begin{equation}
@@
 \end{equation}
+
 \begin{equation}
@@
 The sky distances between the transient and galaxy centroids are calculated as follows, and include the cos-dec factor and a conversion from units of degrees to arcseconds:
+
 \begin{equation}
@@
 Finally, the separation distance $R$ in arcseconds is calculated as:
+
 \begin{equation}
@@
 \end{equation}
+
+\clearpage
+\section{Appendix B: Overcoming Background Galaxy Contamination}\label{sec:appB}
+
+In this appendix, the probability of failed host galaxy associations for nearby transients with large host offsets due to interloping background galaxies is quantified, and used to evaluate the minimum number of potential hosts that should be stored in a {\tt DIAObect} record.
+For simplicity, this appendix uses the effective radius as the separation distance (Section \ref{ssec:options_Re}).
+
+The 10-year {\tt Object} catalogs will include $\sim$4 billion galaxies with $i<25$ mag across the $\sim$20000 $\rm deg^2$ main survey area, known as the ``gold" sample, especially in the context of weak lensing studies \citedsp{LPM-17}.
+However, the 10-year coadded depths will detect galaxies down to 5$\sigma$ limiting magnitudes of 26.1, 27.4, 27.5, 26.8, 26.1, and 24.9 mag in filters {\it ugrizy}; this is $\sim$3 times as many as in the ``gold" sample, or $\sim$10 billion galaxies.
+This high density of background galaxies complicates the process for associating large nearby host galaxies with their transients, especially the rare transients in their outskirts.
+
+Consider a transient at $3R_e$ from the center of a nearby galaxy with $z=0.01$ and an effective radius of $R_e = 10$ kpc ($\sim$50 arcsec).
+In order for this transient to be associated with it's true host, the separation distance for all {\tt Objects} within an area of at least $A_{3R_e} = \pi (3R_e)^2 = 0.0052$ $\rm deg^2$ would need to be considered.
+Based on the final 10-year number of detected galaxies (10 billion) and the total survey area (20000 deg$^2$), that is $\sim (10^10 / 20000) \times 0.0052 \approx 2600$ galaxies.
+Furthermore, the true host galaxy must have a lower separation distance than the $N$ nearest background galaxies, where $N$ is the number of potential hosts that will be listed in the {\tt DIAObject} parameters {\tt potentialHost} and {\tt potentialHostSeparation}.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=8cm]{notebooks/hg_Mr_dist.png}
+\includegraphics[width=8cm]{notebooks/hg_Mr_vs_z.png}
+\includegraphics[width=8cm]{notebooks/hg_rad_dist.png}
+\includegraphics[width=8cm]{notebooks/hg_rad_vs_z.png}
+\includegraphics[width=8cm]{notebooks/hg_z_dist.png}
+\includegraphics[width=8cm]{notebooks/hg_rad_vs_Mr.png}
+\caption{{\it Left:} Histograms of the synthesized absolute $r$-band intrinsic magnitude (top) and radius (middle) for catalog galaxies, and the simulated galaxy redshifts (bottom).
+{\it Right:} Correlation with redshift of the synthesized intrinsic magnitude (top) and radius (middle), and the approximate relation between radius and intrinsic magnitude (bottom).
+\label{fig:simcat}}
+\end{center}
+\end{figure}
+
+To investigate the probability of host-association failure for nearby transients, we simulate a mock catalog of randomly distributed random background interloper galaxies.
+It is based on the same LSST-like mock galaxy catalog used by \citet{2018AJ....155....1G} for studies of photometric redshifts.
+Here, the catalog is limited to galaxies with at least a 5$\sigma$ detection in the {\it griz} bands at the projected 10-year depths listed above.
+The catalog contains redshifts and apparent {\it ugrizy} magnitudes, which are used to approximate intrinsic absolute $r$-band magnitudes.
+The absolute magnitudes are used to synthesize approximate galaxy radii based on the relationship between absolute magnitude and radius for late-type galaxies defined in Figure 3 of \citet{2003MNRAS.343..978S}.
+These synthesized radii are over-estimates because late-types are generally larger than early-types, and because this magnitude-radius relation was defined for SDSS galaxies at lower redshifts than the LSST high-$z$ galaxies it is being applied to.
+This is deliberate, because it will result in upper limits on the rate of interlopers.
+The characteristics of this crude galaxy catalog are illustrated in Figure \ref{fig:simcat}.
+
+Consider again the large nearby galaxy with $R_e=10$ kpc at a redshift of $z=0.01$, for which the sky area within $3R_e$ is $A(3R_e)=0.0052$ $\rm deg^2$ and contains $\sim$2600 background galaxies.
+From the simulated catalog described above, 100 sets of background galaxies are randomly selected.
+For each set, the fraction of the nearby galaxy's area which is covered by the area of interloping background galaxies is calculated: $f_A = \sum{A(3R_{e,{\rm bkg}})} / 0.0052$.
+This fraction is equivalent to the probability that a transient at $3R_e$ from this large nearby host will be within $3R_{e,{\rm bkg}}$ (i.e., closer to) a background interloper.
+The probability that this transient is closer to $N$ interlopers than to its true host is $P_{\rm fail} = (f_A)^N$.
+This is the probability of a failed host association, where ``failed" means that the {\tt DIAObject} record of the $N$ galaxies with the lowest separation distances does not include the true host.
+For this large nearby galaxy, average values of $f_A$ and $P_{\rm fail}$ from the 100 simulated background sets are $f_A = 0.181\pm0.008$ and $P_{\rm fail} (N=3) = 0.006\pm0.0008$
+
+Since this probability of failure is based on the sky density of background galaxies, it is independent of the radius and redshift of the nearby galaxy. 
+However, it does depend on the factor applied to effective radius (i.e., the offset of transient considered) and $N$, the number of potential hosts recorded in the {\tt DIAObject} catalog.
+Figure \ref{fig:pfail} shows the probability of failure as a function of $R_e$ and $N$.
+These are \textbf{upper limits} on the probability of failure, as they are derived from upper-estimates of galaxy radii and the sky density of the final 10-year LSST galaxy catalog.
+
+To assist in the interpretation of Figure \ref{fig:pfail}, the following describes several conclusions drawn from points in this plot:
+\begin{itemize}
+\item Transients with low host offsets, $1R_e$, are closer to (within $1R_{e,{\rm bkg}}$ of) one background interloper $\sim2\%$ of the time (purple point at $1R_e$).
+(However, given the high surface brightness of nearby galaxies, such a background galaxy may well be undetected.)
+\item Transients with high host offsets, $5R_e$, are closer to (within $5R_{e,{\rm bkg}}$ of) one background interloper $\sim50\%$ of the time (purple point at $5R_e$), and closer to six background interlopers $>1\%$ of the time (magenta point at $5R_e$).
+\item In order to achieve a $1\%$ probability of failure for transients offset by $3R_e$ from large nearby galaxies, the {\tt DIAObject} catalog record should include the $N=3$ galaxies with the lowest separation distances (green point at $3R_e$).
+\end{itemize}
+
+Based on Figure \ref{fig:pfail}, in order to reduce the probability of failure in associating nearby, large-offset transients with their true hosts:
+\begin{enumerate}
+\item The separation distances for all galaxies within at least $4R_e \approx 200$ arcsec should be calculated and considered.
+\item The 3 galaxies with the lowest separation distances should be included in the {\tt DIAObject} catalog record.
+\end{enumerate}
+Adopting these recommendations would cause up to $1\%$ of the transients at $3R_e$ from large nearby galaxies to experience a failed host association, where the true host is not listed in the {\tt DIAObject} record.
+Since $3R_e$ encompasses $\sim99\%$ of a galaxy's light, and most transient types are distributed proportional to the light, \textbf{the upper limit} on the host association failure rate for nearby transients should be $\sim0.01\%$.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=12cm]{notebooks/hg_P_vs_Re.png}
+\caption{The probability that a transient will fail to be associated with a large ($R_e=10$ kpc) nearby host galaxy due to background interlopers, as a function of the transient's offset in effective radii from galaxies in the vicinity (including the true host), where ``failure" means the true host's separation distance is not in the top $N$ nearest galaxies.
+\textbf{This is a ``worst case scenario" because it applies to an exceptionally large nearby galaxy, and all background galaxy radii estimates are upper limits} (as described in the text).
+Error bars show the standard deviation from the 100 randomly-generated sets of background galaxies.
+\label{fig:pfail}}
+\end{center}
+\end{figure}
 

### 2020-05-11 draft-hostassoc.tex
v2 updates

@@
 Including the three galaxies (instead of the two or one) with the lowest separation distances is motivated by the high-reward science case of correctly associating transients in the outskirts of large nearby host galaxies, as described in Appendix \ref{sec:appB}.
-This science case also requires that separation distances be initially calculated for all galaxies within $\sim$200 arcsec (i.e., $4R_e$ of a large nearby galaxy); how to identify this sub-set of galaxies in the vicinity (e.g.,.
-radial on-sky distance, HEALpix) is left as a technical implementation detail.
-Large bright nearby galaxies which are not in the {\tt Object} catalogs should have their coordinates and second moments imported so that they are included in this potential host association process.
+This science case also needs separation distances to be initially calculated for all galaxies within $\sim$200 arcsec (i.e., $4R_e$ of a large nearby galaxy).
+How to identify this sub-set of galaxies in the vicinity (e.g.,. radial on-sky distance, HEALpix) is left as a technical implementation detail.
+Furthermore, this science case requires that any large bright nearby galaxies which are not in the {\tt Object} catalogs should have their coordinates and second moments imported so that they are included in this potential host association process -- this is also left as a technical implementation detail.
 
@@
 The separation distance should be calculated with respect to the transient location using the second moments of each {\tt Object}'s luminosity profile, as described in DMTN-XXX.
+Further motivation for these new table elements, as well as specifications that would influence the options for technical implementation, and also described in DMTN-XXX.
 All parameters required for the calculation of the separation distances are already planned to be in the {\tt DIAObject} and {\tt Object} tables, and this change would only add unit64[3] and float[3] per {\tt DIAObject} catalog entry, and per alert.
@@
 
-The sky distances between the transient and galaxy centroids are calculated as follows, and include the cos-dec factor and a conversion from units of degrees to arcseconds:
+The sky distances between the transient and galaxy centroids are calculated as follows, and include the $\cos(\delta)$ factor and a conversion from units of degrees to arcseconds:
 
@@
 
-Finally, the separation distance $R$ in arcseconds is calculated as:
+Finally, the separation distance $R$ in arcseconds is given by:
 
@@
 Consider a transient at $3R_e$ from the center of a nearby galaxy with $z=0.01$ and an effective radius of $R_e = 10$ kpc ($\sim$50 arcsec).
-In order for this transient to be associated with it's true host, the separation distance for all {\tt Objects} within an area of at least $A_{3R_e} = \pi (3R_e)^2 = 0.0052$ $\rm deg^2$ would need to be considered.
+In order for this transient to be associated with it's true host, the separation distance for all {\tt Objects} within a radius of at least $3R_e$, and thus an area of at least $A_{3R_e} = \pi (3R_e)^2 = 0.0052$ $\rm deg^2$, would need to be considered.
 Based on the final 10-year number of detected galaxies (10 billion) and the total survey area (20000 deg$^2$), that is $\sim (10^10 / 20000) \times 0.0052 \approx 2600$ galaxies.
@@
 It is based on the same LSST-like mock galaxy catalog used by \citet{2018AJ....155....1G} for studies of photometric redshifts.
-Here, the catalog is limited to galaxies with at least a 5$\sigma$ detection in the {\it griz} bands at the projected 10-year depths listed above.
-The catalog contains redshifts and apparent {\it ugrizy} magnitudes, which are used to approximate intrinsic absolute $r$-band magnitudes.
+For this experiment the catalog is limited to galaxies with at least a 5$\sigma$ detection in the {\it griz} bands at the projected 10-year depths listed above.
+The catalog contains redshifts and apparent {\it ugrizy} magnitudes, which are used to approximate intrinsic absolute $r$-band magnitudes (without $K$-correction, simply using a distance modulus based on redshift and $M_r=m_r-\mu$).
 The absolute magnitudes are used to synthesize approximate galaxy radii based on the relationship between absolute magnitude and radius for late-type galaxies defined in Figure 3 of \citet{2003MNRAS.343..978S}.
 These synthesized radii are over-estimates because late-types are generally larger than early-types, and because this magnitude-radius relation was defined for SDSS galaxies at lower redshifts than the LSST high-$z$ galaxies it is being applied to.
-This is deliberate, because it will result in upper limits on the rate of interlopers.
+This is a deliberate choice to overestimate, because it will result in upper limits on the rate of interlopers, which will allow for conservative estimates about the effect of background interlopers.
 The characteristics of this crude galaxy catalog are illustrated in Figure \ref{fig:simcat}.
@@
 For each set, the fraction of the nearby galaxy's area which is covered by the area of interloping background galaxies is calculated: $f_A = \sum{A(3R_{e,{\rm bkg}})} / 0.0052$.
-This fraction is equivalent to the probability that a transient at $3R_e$ from this large nearby host will be within $3R_{e,{\rm bkg}}$ (i.e., closer to) a background interloper.
+This fraction is equivalent to the probability that a transient at $3R_e$ from this large nearby host will be within $3R_{e,{\rm bkg}}$ (i.e., will be closer to) a background interloper.
 The probability that this transient is closer to $N$ interlopers than to its true host is $P_{\rm fail} = (f_A)^N$.
@@
 Figure \ref{fig:pfail} shows the probability of failure as a function of $R_e$ and $N$.
-These are \textbf{upper limits} on the probability of failure, as they are derived from upper-estimates of galaxy radii and the sky density of the final 10-year LSST galaxy catalog.
+\textbf{These are upper limits on the probability of failure} because they are derived from upper-estimates of galaxy radii and the sky density of the final 10-year LSST galaxy catalog, as described above.
 
@@
 \item Transients with low host offsets, $1R_e$, are closer to (within $1R_{e,{\rm bkg}}$ of) one background interloper $\sim2\%$ of the time (purple point at $1R_e$).
-(However, given the high surface brightness of nearby galaxies, such a background galaxy may well be undetected.)
+(However, given the high surface brightness of nearby galaxies, such a background galaxy might not be detected.)
 \item Transients with high host offsets, $5R_e$, are closer to (within $5R_{e,{\rm bkg}}$ of) one background interloper $\sim50\%$ of the time (purple point at $5R_e$), and closer to six background interlopers $>1\%$ of the time (magenta point at $5R_e$).
@@
 
-Based on Figure \ref{fig:pfail}, in order to reduce the probability of failure in associating nearby, large-offset transients with their true hosts:
+All of the above experiment can be summarized in two main recommendations aimed at reducing the probability of failure in associating nearby, large-offset transients with their true hosts:
 \begin{enumerate}
@@
 Adopting these recommendations would cause up to $1\%$ of the transients at $3R_e$ from large nearby galaxies to experience a failed host association, where the true host is not listed in the {\tt DIAObject} record.
-Since $3R_e$ encompasses $\sim99\%$ of a galaxy's light, and most transient types are distributed proportional to the light, \textbf{the upper limit} on the host association failure rate for nearby transients should be $\sim0.01\%$.
+Since $3R_e$ encompasses $\sim99\%$ of a galaxy's light, and most transient types are distributed proportional to the light, \textbf{the upper limit on the host association failure rate} for nearby transients should be $\sim0.01\%$.
 
@@
 \caption{The probability that a transient will fail to be associated with a large ($R_e=10$ kpc) nearby host galaxy due to background interlopers, as a function of the transient's offset in effective radii from galaxies in the vicinity (including the true host), where ``failure" means the true host's separation distance is not in the top $N$ nearest galaxies.
-\textbf{This is a ``worst case scenario" because it applies to an exceptionally large nearby galaxy, and all background galaxy radii estimates are upper limits} (as described in the text).
+\textbf{This is a ``worst case scenario" because it applies to an very large nearby galaxy, and all background galaxy radii estimates are upper limits} (as described in the text).
 Error bars show the standard deviation from the 100 randomly-generated sets of background galaxies.
